---
title: "Data preprocessing of gut microbiota cohorts"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: true
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| include: false
#| label: setup

library(readr)
library(biomformat) # BiocManager::install("biomformat")
library(stringr)
library(remotes)
library(phyloseq)
library(MicrobeDS) # remotes::install_github("twbattaglia/MicrobeDS")
library(readxl)
library(tibble)
library(patchwork)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(gt)
library(vegan)
library(ggpubr)
library(seqtime) # remotes::install_github("hallucigenia-sparsa/seqtime")
library(beemStatic) # remotes::install_github('lch14forever/beem-static')
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
library(tidySummarizedExperiment) # BiocManager::install("tidySummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

tmp <- fs::dir_map("R", source)
rm(tmp)

theme_set(theme_light())

```

In this first section, the relevant datasets are imported from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI) as well as the healthy cohorts from the Human Microbiome Project (HMPv13 and HMPv35), NIBIOHN and MORINAGA cohorts, and a "diet intervention" cohort as well. The datasets are available at these following links:

* For **dysbiotic cohorts** (in `01-dysbiotic-cohorts.qmd`):

  - IBD cohort: [https://ibdmdb.org/results](https://ibdmdb.org/results)
  - IBS cohort: [https://data.mendeley.com/datasets/29n2z5r5ph/3](https://data.mendeley.com/datasets/29n2z5r5ph/3)
  - CRC cohort: [https://www.nature.com/articles/s41591-019-0458-7](https://www.nature.com/articles/s41591-019-0458-7)
  - CDI cohort: [https://github.com/grp-bork/cdifficile_Ferretti_2022/tree/main/data/CDI](https://github.com/grp-bork/cdifficile_Ferretti_2022/tree/main/data/CDI)
  
* For **healthy cohorts** (in `02-healthy-cohorts.qmd`):

  - HMPv13 and HMPv35: [https://github.com/twbattaglia/MicrobeDS/tree/master/data](https://github.com/twbattaglia/MicrobeDS/tree/master/data)
  - NIBIOHN and MORINAGA cohorts: [https://link.springer.com/article/10.1186/s12866-021-02215-0](https://link.springer.com/article/10.1186/s12866-021-02215-0)
  - Diet intervention cohort: [https://pubmed.ncbi.nlm.nih.gov/36079725/](https://pubmed.ncbi.nlm.nih.gov/36079725/)
  
Supplementary cohorts will also be used in the analyses, and they will be imported in the third section (see `03-supplementary-cohorts.qmd`):

- IBD cohort without time series (Fransoza et al. 2019): [https://www.nature.com/articles/s41564-018-0306-4](https://www.nature.com/articles/s41564-018-0306-4)
- IBD cohort with time series (Halfvarson et al., 2017): [https://www.nature.com/articles/nmicrobiol20174](https://www.nature.com/articles/nmicrobiol20174)

The final output will be a MAE (MultiAssayExperiment) object, containing data that has been cleaned and rarefied for the downstream analyses.

# Dysbiotic cohorts

```{r 00-data-preprocessing, child = '01-dysbiosis-cohorts.qmd'}
#| eval: true
```

# Healthy cohorts

```{r 00-data-preprocessing, child = '02-healthy-cohorts.qmd'}
#| eval: true
```

# Supplementary cohorts

```{r 00-data-preprocessing, child = '03-supplementary-cohorts.qmd'}
#| eval: false
```

# MAE integration

```{r 00-data-preprocessing, child = '04-MAE-summary.qmd'}
#| eval: true
```

# Saving the preprocessed MAE

We save the preprocessed MAE object as an RDS file for later use in the downstream analyses.

```{r}

saveRDS(mae, file.path(data_dir(), "02-processed-data", str_c("MAE_processed_", today() |> str_remove_all("-"), ".rds")))

```

