---
title: "Data preprocessing for dysbiotic cohorts"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

In this section, we will first import the relevant datasets from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI), and then perform several descriptive analyses for each of those datasets. The analyses will include: data cleaning, taxonomic aggregation (if necessary) and rarefaction. 

## IBD dataset

The inflammatory bowel disease (IBD) dataset was obtained from the [IBDMDB](https://ibdmdb.org/), with the results published in (cite). The authors of the Science paper did not provide information to which study name the IBD dataset comes from, so we choose the HMP2 study, which seems like a reasonable choice since it is the most recent study on this page. For the analyses, we have the following datasets at our disposal:

- `hmp2_metadata.csv`: contains the metadata for the samples.

From Metagenomics (MGX) and products sections, we have the following file available:

- `taxonomic_profiles.biom.gz`: contains the taxonomic profiles and relative abundance data for **stool samples**, in BIOM format.

In the chunks below, we import the metadata and taxonomic profile for the IBD dataset.

```{r}
# Importing the metadata

metadata <- read_csv(file.path(data_dir(), "01-raw-data/01-IBD", "hmp2_metadata_2018-08-20.csv"), col_types = cols(.default = "c"))

# Importing the taxonomic profiles

data <- read_biom(file.path(data_dir(), "01-raw-data/01-IBD", "taxonomic_profiles.biom"))

```

We transform the BIOM object into a `phyloseq` objects in the next steps. For this we need:

- abundance data, a matrix of relative abundance with samples as rows and features (taxa) as columns.
- taxonomy data, a data frame with features (taxa) as rows and taxonomic levels as columns.
- samples data, a data frame with samples as rows and metadata variables as columns.

For abundance data, we need to filter the taxonomic profiles to keep only the bacterial species relevant for our analyses, and to remove the taxonomic levels from the taxonomic names by keeping only the last available level.

```{r}

df <- data |> biom_data() |> as.matrix() |> t() |> as.data.frame() 
abundance <- df |> select(contains("k__Bacteria"))

names(abundance) <- str_extract(names(abundance), "[^|]+$")

```

For taxonomy data, we need to extract the taxonomic levels and names from the taxonomic profiles, and to keep only the last available level for each feature, saved as `taxon_names`, with a letter indicating the taxonomic level between parentheses (e.g., "g" for genus, "f" for family, etc.). Additionally, we also filter the dataset to only keep the "Bacteria" kingdom, since we are only interested in bacterial taxa for our analyses.

```{r}

taxonomy <- 
  tibble(full_name = names(df)) |>
  mutate(
    kingdom = str_extract(full_name, "k__[^|]+"),
    phylum = str_extract(full_name, "p__[^|]+"),
    class = str_extract(full_name, "c__[^|]+"),
    order = str_extract(full_name, "o__[^|]+"),
    family = str_extract(full_name, "f__[^|]+"),
    genus = str_extract(full_name, "g__[^|]+"),
    species = str_extract(full_name, "s__[^|]+")
  ) |>
  mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, kingdom
    )
  ) |>
  mutate(
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(kingdom, "k__"), " (k)")
    )
  ) |>
  filter(kingdom == "k__Bacteria") |>
  select(-full_name)

```

::: {.callout-note}
In the supplementary code provided, the authors filtered out taxa with `_unclassified` annotation. Also, they are also `_nonname` taxa, but they did not filter those out. I filter out both `_unclassified` and `_noname` taxa, but I am not sure if this is the right choice. *Should we filter out both `_unclassified` and `_noname` taxa, or only `_unclassified` taxa?*
:::

We document below how many features have `_unclassified` or `_noname` annotation at each taxonomic level.

```{r}

taxonomy |> 
  summarise(
    kingdom_unclassified = sum(str_detect(kingdom, "_unclassified"), na.rm = TRUE),
    kingdom_noname = sum(str_detect(kingdom, "_noname"), na.rm = TRUE),
    phylum_unclassified = sum(str_detect(phylum, "_unclassified"), na.rm = TRUE),
    phylum_noname = sum(str_detect(phylum, "_noname"), na.rm = TRUE),
    class_unclassified = sum(str_detect(class, "_unclassified"), na.rm = TRUE),
    class_noname = sum(str_detect(class, "_noname"), na.rm = TRUE),
    order_unclassified = sum(str_detect(order, "_unclassified"), na.rm = TRUE),
    order_noname = sum(str_detect(order, "_noname"), na.rm = TRUE),
    family_unclassified = sum(str_detect(family, "_unclassified"), na.rm = TRUE),
    family_noname = sum(str_detect(family, "_noname"), na.rm = TRUE),
    genus_unclassified = sum(str_detect(genus, "_unclassified"), na.rm = TRUE),
    genus_noname = sum(str_detect(genus, "_noname"), na.rm = TRUE),
    species_unclassified = sum(str_detect(species, "_unclassified"), na.rm = TRUE),
    species_noname = sum(str_detect(species, "_noname"), na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "taxonomic_level", values_to = "n") |>
  gt(caption = "Number of features with _unclassified or _noname annotation at each taxonomic level") |>
  tab_options(
    container.height = px(400)
  )

```


For taxa samples, several taxa have the same taxonomic annotation, so we need to aggregate the abundance data by taxonomic annotation, and to keep only one row for each taxon in the taxonomy data.

The duplicates are documented below, with the taxonomic annotation and the number of features with this annotation.

```{r}

taxonomy |> group_by(taxon_name) |> summarise(n = n()) |> filter(n > 1) |>
  gt(caption = "Taxonomic annotation with duplicates") |>
  tab_options(
    container.height = px(400)
  )

```

Since abundance data is in relative scale, we can simply sum the relative abundance of features with the same taxonomic annotation to get the relative abundance of the taxon.

```{r}

abundance <-
  abundance |>
  t() |>
  as.data.frame() |>
  bind_cols(taxonomy |> select(taxon)) |>
  group_by(taxon) |>
  summarise(across(everything(), sum)) |>
  column_to_rownames("taxon") |>
  t() |>
  as.data.frame()

abundance |> head() |> select(1:5)

```

We also need to filter the taxonomy data to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <- 
  taxonomy |> 
  distinct(taxon, .keep_all = TRUE) |> 
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

Since the IBD dataset contains higher level taxonomic annotations in addition to species level annotations, we can also create a `phyloseq` object with only species level annotations, by filtering the taxonomy data to keep only species level annotations, and modify the abundance data accordingly.

Additionally, we also filter out species for which `_unclassified` or `_noname` annotation is present.

```{r}

taxonomy <- 
  taxonomy |> 
  rownames_to_column("taxon") |>
  filter(!is.na(species)) |> 
  column_to_rownames("taxon") |>
  filter(!str_detect(species, "_unclassified|_noname"))

```


```{r}

abundance <- 
  abundance |>
  t() |>
  as.data.frame() |>
  rownames_to_column("taxon") |>
  filter(taxon %in% rownames(taxonomy)) |> 
  column_to_rownames("taxon") |>
  mutate(across(everything(), as.numeric)) |>
  t() |>
  as.data.frame()

```

We check if samples rows sum to 1.

```{r}

row_sums <- rowSums(abundance)

if (all(row_sums == 1)) {
  cat("All samples rows sum to 1.\n")
} else {
  cat("Not all samples rows sum to 1.\n")
}

row_sums |> summary()

```

We renormalize the abundance data to make sure that all samples rows sum to 1 and discard samples for which the sum of relative abundance is 0 (i.e., samples for which we do not have any bacterial species annotation after filtering out unclassified and noname species).

```{r}

dropped <- rownames(abundance)[rowSums(abundance) == 0]
cat("Dropped samples:\n"); print(dropped)

abundance <- abundance[rowSums(abundance) != 0, ] |> (\(x) x / rowSums(x))()

rowSums(abundance) |> summary()

```

For samples data, we need to filter the metadata to keep only the samples for which we have metagenomics data, and to set the sample IDs as row names. Additionally, since not all covariates in the metadata are relevant for our analyses, we also select only a subset of the metadata variables to keep in the samples data. The selected variables are: "Project", "External ID", "Participant ID", "data_type", "week_num", "visit_num", "IntervalName", and "diagnosis".

```{r}

samples <- 
  metadata |> 
  select("Project", "External ID", "Participant ID", "data_type", "week_num", "visit_num", "IntervalName", "diagnosis") |>
  clean_names() |>
  filter(data_type == "metagenomics") |>
  column_to_rownames("external_id")

samples |> glimpse()

```

We also filter out the dropped samples (for which the sum of relative abundance is 0) from the samples data.

```{r}

samples <- 
  samples |> 
  rownames_to_column("sample_id") |>
  filter(!sample_id %in% dropped) |>
  column_to_rownames("sample_id")

```

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibd <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibd

```

## IBS dataset

The irritable bowel syndrome (IBS) dataset was obtained from [Mendeley data](https://data.mendeley.com/datasets/29n2z5r5ph/3) plateform. For the analyses, we have the following datasets at our disposal:

- `metagenomics.xlsx`: contains the metagenomics count data and metadata for biopsy samples, in CSV format. This correspond to the file "Supplementary data I Microbiome data tables stool and biopsy.xlsx" in the plateform (version 3 from 2020-03-11).

In the chunk below, we import the metadata and abundance data for the IBS dataset.

```{r}

data_biopsy <- readxl::read_excel(file.path(data_dir(), "01-raw-data/02-IBS", "metagenomics.xlsx"), sheet = 2)
metadata <- readxl::read_excel(file.path(data_dir(), "01-raw-data/02-IBS", "metagenomics.xlsx"), sheet = 5)

```

We will transform the abundance data into a `phyloseq` object in the next steps, but before that, we need to perform some data cleaning. For this, we need to filter the abundance data to keep only bacterial species, and to remove the taxonomic levels from the taxonomic names by keeping only the last available level.

```{r}

abundance <- data_biopsy |> filter(str_detect(`#OTU ID`, "k__Bacteria"))

```

For taxonomy data, we need to extract the taxonomic levels and names from the taxonomic profiles (OTU column in `data_biopsy`), and to keep only the last available level for each feature, saved as `taxon_names`, with a letter indicating the taxonomic level between parentheses (e.g., "g" for genus, "f" for family, etc.).

```{r}

taxonomy <-
  tibble(full_name = abundance$`#OTU ID`) |>
  mutate(
    kingdom = str_extract(full_name, "k__[^;]+"),
    phylum = str_extract(full_name, "p__[^;]+"),
    class = str_extract(full_name, "c__[^;]+"),
    order = str_extract(full_name, "o__[^;]+"),
    family = str_extract(full_name, "f__[^;]+"),
    genus = str_extract(full_name, "g__[^;]+"),
    species = str_extract(full_name, "s__[^;]+")
  ) |>
    mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, kingdom
    )
  ) |>
  mutate(
    across(c(species, genus, family, order, class, phylum, kingdom), 
           ~ str_remove_all(., "\\[|\\]")),
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(kingdom, "k__"), " (k)")
    )
  ) |>
  select(-full_name)

```

We check if there are any missing species annotations.

```{r}

cat("There are", nrow(taxonomy |> filter(is.na(species)) == "TRUE"), "features without species annotation.\n")

```

We additionally check if there are any duplicated taxonomic annotations, and if so, we document them by showing the taxonomic annotation and the number of features with this annotation. The duplicates will be merged in taxonomy, as well as in abundance data.

```{r}

taxonomy |> 
  group_by(taxon_name) |> 
  summarise(n = n()) |> 
  filter(n > 1) |>
  gt(caption = "Taxonomic annotation with duplicates")

```

To merge duplicates in abundance, we can simply sum the counts of features with the same taxonomic annotation to get the count of the taxon.

```{r}

abundance <-
  abundance |>
  bind_cols(taxonomy |> select(taxon)) |>
  group_by(taxon) |>
  summarise(across(-`#OTU ID`, sum)) |>
  column_to_rownames("taxon")

abundance |> head() |> select(1:5)

```

We filter duplicates in taxonomy to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <- 
  taxonomy |> 
  distinct(taxon, .keep_all = TRUE) |> 
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

For sample (metadata) data, we only keep relevant metadata variables for our analyses, and we set the sample IDs as row names. The selected variables are: 

- `SampleID`: the sample ID, which will be set as row names.
- `study_id`: the participant ID, which can be used to link samples from the same participant.
- `Timepoint`: the time points of the sample collection. Originally encoded as "1st" and "2nd", we will rename them "1" and "2" for easier handling.
- `Cohort`: the cohort to which the sample belongs. It consists of three levels: "H" for healthy controls, "D" for IBS patients for which the stool form was classified as predominantly diarrhea, and "C" for IBS patients for which the stool form was classified as predominantly constipation. Since we work only with biopsy samples, "C" and "D" levels will be merged into a single level "IBS" for the analyses.

```{r}

samples <-
  metadata |>
  clean_names() |>
  select(sample_id, study_id, timepoint, cohort) |>
  mutate(cohort = ifelse(cohort %in% c("C", "D"), "IBS", cohort),
         cohort = cohort |> factor()) |>
  mutate(timepoint = str_replace_all(timepoint, c("1st" = "1", "2nd" = "2")),
         timepoint = timepoint |> factor()) |>
  column_to_rownames("sample_id")

samples |> glimpse()

```
We will then rarefy the IBS dataset, so that all samples have the same sequencing depth, and filter out low-abundance taxa. We will use the `rrarefy` function from the `vegan` package to perform rarefaction.

First, we separate the dataset into two groups: "healthy" and "diseased" so that rarefaction (and prevalence filtering) is performed separately for the two groups, to avoid biasing the results by removing more samples from one group than the other.

```{r}

tmp <-
  abundance |>
  rownames_to_column("taxon") |>
  pivot_longer(-taxon, names_to = "sample_id", values_to = "count") |>
  left_join(
    samples |> rownames_to_column("sample_id") |> select(sample_id, cohort)
  ) |>
  mutate(
    status = ifelse(cohort == "H", "Healthy", "Diseased")
  )

```


```{r}

tmp_healthy <- tmp |> filter(status == "Healthy")
tmp_diseased <- tmp |> filter(status == "Diseased")

```

We show the distribution of sequencing depth across samples, to choose a rarefaction depth.

::: {.callout-note}
Note that a classical rarefaction threshold would be to simply take the median depth, but in this case, we choose to take the median depth divided by 2, to keep more samples in the dataset, especially given the relatively small number of samples in the IBS dataset.
:::

```{r}

tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across healthy samples")

tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across diseased samples")

```

We will rarefy at a depth of median(depth)/2; this is done in the chucks below, where we also filters out samples with depth < median(depth)/2.

```{r}

# for healthy samples

tmp_depth <- tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_healthy <- tmp_healthy |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```


```{r}

# for diseased samples

tmp_depth <- tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_diseased <- tmp_diseased |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```

We convert the rarefied data back to wide format, and we merge the healthy and diseased datasets back together.

```{r}

abundance <- bind_rows(tmp_rarefied_healthy, tmp_rarefied_diseased) |>
  select(sample_id, taxon, count) |>
  pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
  column_to_rownames("sample_id")

abundance |> head() |> select(1:5)

```

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibs <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibs

```


## CRC dataset

The colorectal cancer (CRC) dataset was obtained from the DDBJ Sequence Read Archive (DRA) under the names [DRA006684](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA006684) and [DRA008156](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA008156); however, those are raw sequencing data, and the authors mentioned having used only processed datasets.

Processed datasets seems to be available as [Supplementary Tables](https://www.nature.com/articles/s41591-019-0458-7#Sec23) in (cite), so I will use those for the analyses. Therefore, we have the following datasets at our disposal:

- `41591_2019_458_MOESM3_ESM.xlsx`: contains the processed datasets for stool samples, in Excel format. This correspond to the file "Supplementary data 3.xlsx" in the article (version 1 from 2019-11-28).

We import those datasets in the chunks below.

```{r}

data <- read_xlsx(file.path(data_dir(), "01-raw-data/03-CRC", "41591_2019_458_MOESM3_ESM.xlsx"), skip = 3, sheet = 13)

metadata <- read_xlsx(file.path(data_dir(), "01-raw-data/03-CRC", "41591_2019_458_MOESM3_ESM.xlsx"), skip = 2, sheet = 4)

```

As in the previous datasets, we need to perform some data cleaning before creating the `phyloseq` object. For abundance data, we need to filter the dataset to keep only bacterial species.

First, we notice that the number of samples in metadata and abundance data do not match, so we need to filter the abundance data to keep only samples for which we have metadata.

```{r}

abundance <- data |> select(any_of(metadata$Subject_ID))

```

Now, we see that we have `r nrow(metadata)`samples in metadata, and `r ncol(abundance)` samples in abundance data, so there is still a mismatch between the two datasets. We document this below.

```{r}

setdiff(metadata$Subject_ID, colnames(abundance))

```
After checking manually, we see that this sample is present in the metadata, but not in the abundance data, so we can simply filter it out from the metadata.

```{r}

metadata <- metadata |> filter(!Subject_ID %in% setdiff(metadata$Subject_ID, colnames(abundance)))

```

We create the taxonomy data in the next few steps.

::: {.callout-note}
We see here that we do not have the complete taxonomy lineage for the features in the abundance data, but only the last available taxonomic level (e.g., species). This will make it difficult to filter the dataset to keep only bacterial species, since we do not have the taxonomic annotation for the kingdom level. *Can we assume that all species are bacteria and go ahead with the analyses?*
:::

We first create the taxonomy table by extracting the taxonomic annotation from the feature names in the abundance data, and by keeping only the last available level for each feature, saved as `taxon_names`.

```{r}

taxonomy <- 
  tibble(full_name =  data[1] |> pull()) |>
  filter(!str_detect(full_name, "-1")) |>
  mutate(
    taxon_name = full_name,
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = str_c("g_", word(taxon_name, 1)),
    species = case_when(
      word(taxon_name, 2) == "sp." ~ NA,
      is.na(word(taxon_name, 2)) ~ NA,
      TRUE ~ str_c("s_", word(taxon_name, 2)),
    ),
    taxon = taxon_name
  ) |>
  select(kingdom, phylum, class, order, family, genus, species, taxon_name, taxon)

```

We check if there are duplicates in taxa species, and document them below.

```{r}

taxonomy |> 
  group_by(taxon_name) |> 
  summarise(n = n()) |> 
  filter(n > 1) |>
  gt(caption = "Taxonomic annotation with duplicates")

```

No duplicates are found in the taxonomic annotation, so we can keep all features in the abundance data, and we can simply set the taxonomic annotation as row names for the taxonomy data.

```{r}

taxonomy <- taxonomy |> column_to_rownames("taxon")

taxonomy |> glimpse()

```

For abundance data, we only need to set the taxonomic annotation as row names and filter the dataset to keep only taxa with taxonomic annotation in the taxonomy data.

```{r}

abundance <- data |> column_to_rownames(var = names(data)[1])

abundance <- abundance |> filter(rownames(abundance) %in% rownames(taxonomy))

```

For sample (metadata) data, we only keep relevant metadata variables for our analyses, and we set the sample IDs as row names. The selected variables are:

- `Subject_ID`: the sample ID, which will be set as row names.
- `Group`: the health status of the samples.

We document the different levels of the variable `Group` below.

```{r}

metadata |> 
  clean_names() |> 
  select(subject_id, group) |>
  group_by(group) |> 
  summarise(n = n()) |>
  gt(caption = "Health status of samples")

```

They classified the samples into nine groups according to colonoscopic and histological findings: 

- normal (no remarkable colonoscopic findings) and a few polyps (up to two small (<5 mm) polyps) were classified as "Healthy";
- multiple polypoid adenomas with low-grade dysplasia and polypoid adenoma(s) with high-grade dysplasia were classified as "MP";
- stage 0 CRC was classified as "Stage_0"; 
- stage I CRC and stage II CRC were classified as "Stage_I_II"; 
- stage III CRC and stage IV CRC were classified as "Stage_III_IV". 
- the remaining group was normal with a history of colorectal surgery, and was classified as "HS".

We filter the metadata to keep only the selected variables, and set the sample IDs as row names.

```{r}

samples <- 
  metadata |> 
  clean_names() |> 
  select(subject_id, group) |>
  column_to_rownames("subject_id") |>
  mutate(group = group |> as.factor()) |>
  mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ NA_character_
  ))

```

We filter samples for which we do not have a clear health status (i.e., samples with "HS" group).

```{r}

samples <- samples |> filter(!is.na(cohort)) 

samples |> glimpse()

```

We also filter abundance data to keep only samples for which we have a clear health status in the metadata.

```{r}

abundance <- abundance |> select(any_of(rownames(samples)))

abundance |> head() |> select(1:5)

```

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_crc <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_crc

```


## CDI dataset

The *Clostridium difficile* infection (CDI) metagenomics dataset was obtained from the [Github repository](https://github.com/grp-bork/cdifficile_Ferretti_2022/tree/main/data/CDI) associated with the original article (cite). [indicate the date of the latest version for these datasets]. For the analysis, we have the following datasets at our disposal: 

- `CDI_mergedControl_samples.csv`: contains metadata for both control and CDI samples, in CSV format.
- `mOTU20_profiles_filteredCDI.csv`: contains abundance count data for stool samples, in CSV format.
- `motus_taxonomy.csv`: contains the taxonomic annotation for the features in the abundance data, in CSV format.

In the chunks below, we will import the metadata, abundance data, and taxonomic annotation for the CDI dataset.

```{r}

metadata <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "CDI_mergedControl_samples.csv"))
abundance <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "mOTU20_profiles_filteredCDI.csv"))
taxonomy <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "motus_taxonomy.csv"))

```

::: {.callout-note}
We see here that we do not have the complete taxonomy lineage for the features in the abundance data, but only the last available taxonomic level (e.g., species). This will make it difficult to filter the dataset to keep only bacterial species, since we do not have the taxonomic annotation for the kingdom level. *Can we assume that all species are bacteria and go ahead with the analyses?*
:::

First, we need to filter taxonomy data to merge taxon with the same taxonomic annotation, and to keep only one row for each taxon, with the taxonomic annotation as row names. Additionally, we fill the taxonomic levels based on the taxonomic annotation.

```{r}

taxonomy <- 
  taxonomy |>
  mutate(
    taxon_name = coalesce(
      str_remove(mOTU_name, "\\[.*\\]")),
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = str_c("g_", word(taxon_name, 1)),
    species = case_when(
      word(taxon_name, 2) == "sp." ~ NA,
      is.na(word(taxon_name, 2)) ~ NA,
      TRUE ~ str_c("s_", word(taxon_name, 2))
    ),
    taxon = str_c(taxon_name)
  ) |>
  select(kingdom, phylum, class, order, family, genus, species, taxon, taxon_name)

```

We now filter the abundance data by first replacing the feature names in the abundance data with the taxonomic annotation from the taxonomy data, and then summing the counts of features with the same taxonomic annotation to get the count of the taxon.

```{r}

abundance <-
  abundance |>
  filter(X1 != "-1") |>
  column_to_rownames(var = names(abundance)[1]) |>
  as.data.frame() |>
  bind_cols(taxonomy |> select(taxon_name)) |>
  group_by(taxon_name) |>
  summarise(across(everything(), sum)) |>
  column_to_rownames("taxon_name") |>
  as.data.frame()

```

We filter the taxonomy data to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <-
  taxonomy |>
  distinct(taxon_name, .keep_all = TRUE) |>
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

We will now filter the metadata to keep only relevant variables for our analyses, and to set the sample IDs as row names. The selected variables are: 

- `Samples`: the sample ID, which will be set as row names.
- `Gut_sample_type`: the type of the sample, here only "stool" samples are available.
- `Status`: the health status of the sample, with two levels: "Healthy" and "Diseased".
- `subject_disease_status`: gives further information about the health status of the samples.

```{r}

samples <-
  metadata |>
  clean_names() |>
  select(samples, gut_sample_type, status, subject_disease_status) |>
  column_to_rownames("samples")

```

We document the variable `subject_disease_status` to see how many samples have been diagnosed with other gut-related diseases.

```{r}

samples |>
  group_by(status) |>
  summarise(n = n()) |>
  gt(caption = "Number of healthy and diseased samples")

```


```{r}

samples |>
  group_by(status, subject_disease_status) |>
  summarise(n = n()) |>
  gt(caption = "Health status of healthy and diseased samples")

```

::: {.callout-warning}
Based on the documentation of the variable `subject_disease_status`, we see that all of the "Healthy" samples are control and none of them have a gut-related disease. However, among "Diseased" samples, only 202 among the 348 diseased samples have been diagnosed with CDI. *Do we keep all the "Diseased" samples for our analyses, or do we only keep the 202 samples diagnosed with CDI?*.
:::

I filter the dataset to keep only the 202 samples diagnosed with CDI, as well as the healthy samples (CTR).

::: {.callout-note}
I could have also kept some other diseased samples, such as those "Asymptotically colonized", but there are only four of them.
:::

```{r}

samples <- samples |> filter(subject_disease_status == "CDI" | subject_disease_status == "CTR")

samples |> glimpse()

```

We filter the abundance data to keep only the samples for which we have the corresponding filtered metadata.

```{r}

abundance <- abundance |> 
  select(any_of(rownames(samples)))

abundance |> head() |> select(1:5)

```

We will then rarefy the CDI dataset, so that all samples have the same sequencing depth, and filter out low-abundance taxa. We will use the `rrarefy` function from the `vegan` package to perform rarefaction.

First, we separate the dataset into two groups: "healthy" and "diseased" so that rarefaction (and prevalence filtering) is performed separately for the two groups, to avoid biasing the results by removing more samples from one group than the other.

```{r}

tmp <-
  abundance |>
  rownames_to_column("taxon") |>
  pivot_longer(-taxon, names_to = "sample_id", values_to = "count") |>
  left_join(
    samples |> rownames_to_column("sample_id") |> select(sample_id, status)
  )

```


```{r}

tmp_healthy <- tmp |> filter(status == "Healthy")
tmp_diseased <- tmp |> filter(status == "Diseased")

```

We show the distribution of sequencing depth across samples, to choose a rarefaction depth.

::: {.callout-note}
Note that a classical rarefaction threshold would be to simply take the median depth, but in this case, we choose to take the median depth divided by 2, to keep more samples in the dataset.
:::

```{r}

tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across healthy samples")

tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across diseased samples")

```

We will rarefy at a depth of median(depth)/2; this is done in the chucks below, where we also filters out samples with depth < median(depth)/2.

```{r}

# for healthy samples

tmp_depth <- tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_healthy <- tmp_healthy |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```


```{r}

# for diseased samples

tmp_depth <- tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_diseased <- tmp_diseased |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```

We convert the rarefied data back to wide format, and we merge the healthy and diseased datasets back together.

```{r}

abundance <- bind_rows(tmp_rarefied_healthy, tmp_rarefied_diseased) |>
  select(sample_id, taxon, count) |>
  pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
  column_to_rownames("sample_id")

abundance |> head() |> select(1:5)

```

Since several samples have been removed after rarefaction, we also need to filter the metadata to keep only the samples for which we have abundance data after rarefaction.

```{r}

samples <- samples |> rownames_to_column("sample_id") |> filter(sample_id %in% rownames(abundance)) |> column_to_rownames("sample_id")

```

We finally create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_cdi <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_cdi

```


```{r}
#| include: false

rm(data, data_biopsy, metadata, df, samples, taxonomy, abundance)

```

