---
title: "Multi-assay Experiment (MAE) integration"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

In this section, we will integrate the different cohorts datasets using the `MultiAssayExperiment` (MAE) class from Bioconductor. This class provides a convenient way to manage and analyze the cohorts.

We first transform the datasets into `SummarizedExperiment` objects.

```{r}

# Diseased cohorts

se_ibd <- phyloseq_to_SE(ps_ibd)
se_ibs <- phyloseq_to_SE(ps_ibs)
se_cdi <- phyloseq_to_SE(ps_cdi)
se_crc <- phyloseq_to_SE(ps_crc)

# Healthy cohorts

se_hmpv13 <- phyloseq_to_SE(ps_hmpv13)
se_hmpv35 <- phyloseq_to_SE(ps_hmpv35)
se_NIBIOHN <- phyloseq_to_SE(ps_NIBIOHN)
se_MORINAGA <- phyloseq_to_SE(ps_MORINAGA)

```

We create below a `MultiAssayExperiment` object that contains all the `SummarizedExperiment` objects.

```{r}
# Create a list of the SummarizedExperiment objects

se_list <- list(
  IBD = se_ibd,
  IBS = se_ibs,
  CDI = se_cdi,
  CRC = se_crc,
  HMPV13 = se_hmpv13,
  HMPV35 = se_hmpv35,
  NIBIOHN = se_NIBIOHN,
  MORINAGA = se_MORINAGA
)

```


```{r}

# Create the MultiAssayExperiment object

mae <- MultiAssayExperiment(experiments = se_list)

mae

```
