---
title: "Descriptive analysis of gut microbiota cohorts datasets"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

In this section, we will perform a descriptive analysis of the datasets used in this project.

```{r}
#| include: false

library(readr)
library(biomformat) # BiocManager::install("biomformat")
library(stringr)
library(remotes)
library(phyloseq)
library(MicrobeDS) # remotes::install_github("twbattaglia/MicrobeDS")
library(readxl)
library(tibble)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(gt)
library(vegan)
library(ggpubr)
library(seqtime) # remotes::install_github("hallucigenia-sparsa/seqtime")
library(beemStatic) # remotes::install_github('lch14forever/beem-static')
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
library(tidySummarizedExperiment) # BiocManager::install("tidySummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

tmp <- fs::dir_map("R", source)
rm(tmp)

theme_set(theme_light())

```

## Importing the MAE object

```{r}

mae_file <- 
  fs::dir_ls(str_c(data_dir(), "/02-processed-data"), regexp = "MAE_processed_.*\\.rds$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)


cat(str_c("Loading MultiAssayExperiment from: 00_data/", mae_file |> str_remove(data_dir()), "\n"))

mae <- readRDS(mae_file)

rm(mae_file)

mae

```

## IBD cohort (Franzosa et al., 2018)

In this section, we will apply BEEM-Static to the differents cohorts from the main analysis, and additionnaly compute beta diversity, evenness, and richness measures.

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["IBD"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["IBD"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```

We compute the number of taxa and samples in the dataset, to get an sense of the size of the dataset.

```{r}

n_taxa <- mae[["IBD"]] |> assay() |> nrow()
cat("Number of taxa:", n_taxa, "\n")

n_samples <- mae[["IBD"]] |> assay() |> ncol()
cat("Number of samples:", n_samples, "\n")

```
```{r}

tmp |>
  group_by(participant_id, diagnosis) |>
  summarise(n_samples = n_distinct(sample_id), .groups = "drop") |>
  gt(caption = "Number of samples in the IBD dataset") |>
  tab_options(
    container.height = px(400)
  )

```


```{r}
# separate healthy and IBD samples

mae[["IBD"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  mutate(
    status = case_when(
      diagnosis == "UC" ~ "Diseased",
      diagnosis == "CD" ~ "Diseased",
      TRUE ~ "Healthy"
    ),
    status = factor(status)
  ) |> 
  dplyr::count(status) |>
  gt(caption = "Number of healthy and diseased samples in the IBS dataset")

```

::: {.callout-important}
Rarefaction cannot be applied to the IBD dataset, since counts are not available for this dataset, and only relative abundances are available. 
:::

We will compute beta diversity, evenness, and richness measures on the rarefied dataset. We will use the `vegan` package to compute these metrics.

- Richness = $\sum_{i=1}^{S} I(x_i > 0)$, where $S$ is the total number of taxa and $x_i$ is the abundance of taxon $i$ in the sample.
- Shannon = $-\sum_{i=1}^{S} p_i \log(p_i)$, where $p_i$ is the relative abundance of taxon $i$ in the sample.
- Pielou's evenness = $\frac{-\sum_{i=1}^{S} p_i \log(p_i)}{\log(\sum_{i=1}^{S} I(x_i > 0))}$
- Sheldon evenness = $\frac{exp(-\sum_{i=1}^{S} p_i \log(p_i))}{\sum_{i=1}^{S} I(x_i > 0)}$
- Beta diversity will be computed using Bray-Curtis dissimilarity.

The difference between groups will be assessed using the Wilcoxon rank-sum test.

```{r}

bc_mat <- tmp |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  as.matrix()

```


```{r}

bc_per_sample <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  rowMeans() |>
  as_tibble(rownames = "sample_id") |>
  dplyr::rename(bc = value)

```


```{r}

bc_long <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  as_tibble(rownames = "sample1") |>
  pivot_longer(-sample1, names_to = "sample2", values_to = "bc") |>
  filter(sample1 < sample2)

sample_meta <- mae[["IBD"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  select(sample_id, participant_id, diagnosis)

bc_intra <- bc_long |>
  select(sample1, sample2, bc) |>
  left_join(sample_meta, by = c("sample1" = "sample_id")) |>
  select(sample1, sample2, bc, subject1 = participant_id, cohort1 = diagnosis) |>
  left_join(sample_meta, by = c("sample2" = "sample_id")) |>
  select(sample1, sample2, bc, subject1, cohort1, subject2 = participant_id, cohort2 = diagnosis) |>
  filter(subject1 == subject2) |>
  group_by(subject1, cohort1) |>
  summarise(bc_intra = mean(bc), .groups = "drop") |>
  select(participant_id = subject1, diagnosis = cohort1, bc_intra)

```


```{r}

metrics <- tmp |>
  group_by(sample_id) |>
  summarise(
    richness = sum(rel_ab > 0),
    shannon  = vegan::diversity(rel_ab, index = "shannon"),
    pielou   = shannon / log(richness),
    sheldon  = exp(shannon) / richness,
    .groups  = "drop"
  ) |>
  left_join(bc_per_sample, by = "sample_id") |>
  left_join(
    mae[["IBD"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  ) |>
  group_by(participant_id, diagnosis) |>
  summarise(across(c(richness, shannon, pielou, sheldon, bc), mean), .groups = "drop") |>
  left_join(bc_intra, by = join_by(participant_id, diagnosis)) |>
  mutate(
    status = case_when(
      diagnosis == "UC" ~ "Diseased",
      diagnosis == "CD" ~ "Diseased",
      TRUE ~ "Healthy"
    ),
    status = factor(status)
  )

```


```{r}
# richness plot

metrics |> 
  ggplot() +
  aes(x = status, y = richness, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Richness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# shannon diversity plot

metrics |> 
  ggplot() +
  aes(x = status, y = shannon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Shannon diversity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Pielou evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = pielou, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Pielou evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Sheldon evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = sheldon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Sheldon evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = status, y = bc, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = status, y = bc_intra, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status",
       caption = "Intra-subject Bray-Curtis dissimilarity, \ncomputed as the average Bray-Curtis dissimilarity between samples from the same subject") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```

## IBS cohort (Mars et al., 2020)

We first import the IBS dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["IBS"]] |> 
  assay() |> 
  t() |> 
  as.data.frame() |> 
  rownames_to_column("sample_id") |> 
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["IBS"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, study_id, timepoint, cohort), 
    by = join_by(sample_id)
  )

```


```{r}

tmp |>
  group_by(study_id, cohort) |>
  summarise(n_samples = n_distinct(sample_id), .groups = "drop") |>
  gt(caption = "Number of samples in the IBS dataset, separated by study and cohort") |>
  tab_options(
    container.height = px(400)
  )

```


We compute the number of taxa and samples in the dataset, to get an sense of the size of the dataset.

```{r}

n_taxa <- mae[["IBS"]] |> assay() |> nrow()
cat("Number of taxa:", n_taxa, "\n")

n_samples <- mae[["IBS"]] |> assay() |> ncol()
cat("Number of samples:", n_samples, "\n")

```


```{r}
# separate healthy and IBS samples

mae[["IBS"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  mutate(
    status = case_when(
      cohort == "H" ~ "Healthy",
      cohort == "IBS" ~ "Diseased"
    ),
    status = factor(status)
  ) |> 
  dplyr::count(status) |>
  gt(caption = "Number of healthy and diseased samples in the IBS dataset")

```

We will compute beta diversity, evenness, and richness measures on the rarefied dataset. We will use the `vegan` package to compute these metrics.

- Richness = $\sum_{i=1}^{S} I(x_i > 0)$, where $S$ is the total number of taxa and $x_i$ is the abundance of taxon $i$ in the sample.
- Shannon = $-\sum_{i=1}^{S} p_i \log(p_i)$, where $p_i$ is the relative abundance of taxon $i$ in the sample.
- Pielou's evenness = $\frac{-\sum_{i=1}^{S} p_i \log(p_i)}{\log(\sum_{i=1}^{S} I(x_i > 0))}$
- Sheldon evenness = $\frac{exp(-\sum_{i=1}^{S} p_i \log(p_i))}{\sum_{i=1}^{S} I(x_i > 0)}$
- Beta diversity will be computed using Bray-Curtis dissimilarity.

The difference between groups will be assessed using the Wilcoxon rank-sum test.

```{r}

bc_mat <- tmp |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  as.matrix()

```


```{r}

bc_per_sample <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  rowMeans() |>
  as_tibble(rownames = "sample_id") |>
  dplyr::rename(bc = value)

```


```{r}

bc_long <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  as_tibble(rownames = "sample1") |>
  pivot_longer(-sample1, names_to = "sample2", values_to = "bc") |>
  filter(sample1 < sample2)

sample_meta <- mae[["IBS"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  select(sample_id, study_id, cohort)

bc_intra <- bc_long |>
  select(sample1, sample2, bc) |>
  left_join(sample_meta, by = c("sample1" = "sample_id")) |>
  select(sample1, sample2, bc, subject1 = study_id, cohort1 = cohort) |>
  left_join(sample_meta, by = c("sample2" = "sample_id")) |>
  select(sample1, sample2, bc, subject1, cohort1, subject2 = study_id, cohort2 = cohort) |>
  filter(subject1 == subject2) |>
  group_by(subject1, cohort1) |>
  summarise(bc_intra = mean(bc), .groups = "drop") |>
  select(study_id = subject1, cohort = cohort1, bc_intra)

```


```{r}

metrics <- tmp |>
  group_by(sample_id) |>
  summarise(
    richness = sum(rel_ab > 0),
    shannon  = vegan::diversity(rel_ab, index = "shannon"),
    pielou   = shannon / log(richness),
    sheldon  = exp(shannon) / richness,
    .groups  = "drop"
  ) |>
  left_join(bc_per_sample, by = "sample_id") |>
  left_join(
    mae[["IBS"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |>
      select(sample_id, study_id, timepoint, cohort),
    by = join_by(sample_id)
  ) |>
  group_by(study_id, cohort) |>
  summarise(across(c(richness, shannon, pielou, sheldon, bc), mean), .groups = "drop") |>
  left_join(bc_intra, by = join_by(study_id, cohort)) |>
  mutate(
    status = case_when(
      cohort == "H"   ~ "Healthy",
      cohort == "IBS" ~ "Diseased"
    ),
    status = factor(status)
  )

```


```{r}
# richness plot

metrics |> 
  ggplot() +
  aes(x = status, y = richness, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Richness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# shannon diversity plot

metrics |> 
  ggplot() +
  aes(x = status, y = shannon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Shannon diversity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Pielou evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = pielou, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Pielou evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Sheldon evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = sheldon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Sheldon evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Inter subject Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = status, y = bc, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Intra subject Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = status, y = bc_intra, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status", 
       caption = "Intra-subject Bray-Curtis dissimilarity, \ncomputed as the average Bray-Curtis dissimilarity between samples from the same subject") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```

::: {.callout-note}
All the above metrics for the IBS dataset, computed on rarefied data, do not show a significant difference between healthy and diseased samples. This is in line with conclusion from the paper from Lopez et al. (2026). In the original paper from Mars et al. (2020), the authors calculated Shannon diversity on healthy and diseases samples and found a similar trend (although no p-values between groups were reported in the paper).
:::

## CRC cohort (Ferretti et al., 2022)

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["CRC"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["CRC"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, group, cohort),
    by = join_by(sample_id)
  )

```

We compute the number of taxa and samples in the dataset, to get an sense of the size of the dataset.

```{r}

n_taxa <- mae[["CRC"]] |> assay() |> nrow()
cat("Number of taxa:", n_taxa, "\n")

n_samples <- mae[["CRC"]] |> assay() |> ncol()
cat("Number of samples:", n_samples, "\n")

```


```{r}
# separate healthy and CRC samples

mae[["CRC"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  dplyr::count(cohort) |>
  gt(caption = "Number of healthy and diseased samples in the CRC dataset")

```

::: {.callout-important}
Rarefaction cannot be applied to the IBD dataset, since counts are not available for this dataset, and only relative abundances are available. 
:::

We will compute beta diversity, evenness, and richness measures on the CRC dataset. We will use the `vegan` package to compute these metrics.

- Richness = $\sum_{i=1}^{S} I(x_i > 0)$, where $S$ is the total number of taxa and $x_i$ is the abundance of taxon $i$ in the sample.
- Shannon = $-\sum_{i=1}^{S} p_i \log(p_i)$, where $p_i$ is the relative abundance of taxon $i$ in the sample.
- Pielou's evenness = $\frac{-\sum_{i=1}^{S} p_i \log(p_i)}{\log(\sum_{i=1}^{S} I(x_i > 0))}$
- Sheldon evenness = $\frac{exp(-\sum_{i=1}^{S} p_i \log(p_i))}{\sum_{i=1}^{S} I(x_i > 0)}$
- Beta diversity will be computed using Bray-Curtis dissimilarity.

The difference between groups will be assessed using the Wilcoxon rank-sum test.

```{r}

bc_mat <- tmp |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  as.matrix()

```


```{r}

bc_per_sample <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  rowMeans() |>
  as_tibble(rownames = "sample_id") |>
  dplyr::rename(bc = value)

```


```{r}

metrics <- tmp |>
  group_by(sample_id) |>
  summarise(
    richness = sum(rel_ab > 0),
    shannon  = vegan::diversity(rel_ab, index = "shannon"),
    pielou   = shannon / log(richness),
    sheldon  = exp(shannon) / richness,
    .groups  = "drop"
  ) |>
  left_join(bc_per_sample, by = "sample_id") |>
  left_join(
    mae[["CRC"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, group, cohort),
    by = join_by(sample_id)
  ) |>
  group_by(sample_id, cohort) |>
  summarise(across(c(richness, shannon, pielou, sheldon, bc), mean), .groups = "drop")

```


```{r}
# richness plot

metrics |> 
  ggplot() +
  aes(x = cohort, y = richness, fill = cohort, color = cohort) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Richness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("H", "CRC")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# shannon diversity plot

metrics |> 
  ggplot() +
  aes(x = cohort, y = shannon, fill = cohort, color = cohort) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Shannon diversity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("H", "CRC")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Pielou evenness plot

metrics |> 
  ggplot() +
  aes(x = cohort, y = pielou, fill = cohort, color = cohort) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Pielou evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("H", "CRC")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Sheldon evenness plot

metrics |> 
  ggplot() +
  aes(x = cohort, y = sheldon, fill = cohort, color = cohort) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Sheldon evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("H", "CRC")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}

# Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = cohort, y = bc, fill = cohort, color = cohort) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("H", "CRC")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```

## CDI cohort (Yashida et al., 2020)

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["CDI"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CDI"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, status, subject_disease_status),
    by = join_by(sample_id)
  ) |>
  filter(subject_disease_status == "CDI" | subject_disease_status == "CTR")

```

We compute the number of taxa and samples in the dataset, to get an sense of the size of the dataset.

```{r}

n_taxa <- mae[["CDI"]] |> assay() |> nrow()
cat("Number of taxa:", n_taxa, "\n")

n_samples <- mae[["CDI"]] |> assay() |> ncol()
cat("Number of samples:", n_samples, "\n")

```


```{r}
# separate healthy and CDI samples

mae[["CDI"]] |> 
  colData() |> 
  as.data.frame() |> 
  as_tibble(rownames = "sample_id") |>
  dplyr::count(status) |>
  gt(caption = "Number of healthy and diseased samples in the CDI dataset")

```

We will compute beta diversity, evenness, and richness measures on the rarefied dataset. We will use the `vegan` package to compute these metrics.

- Richness = $\sum_{i=1}^{S} I(x_i > 0)$, where $S$ is the total number of taxa and $x_i$ is the abundance of taxon $i$ in the sample.
- Shannon = $-\sum_{i=1}^{S} p_i \log(p_i)$, where $p_i$ is the relative abundance of taxon $i$ in the sample.
- Pielou's evenness = $\frac{-\sum_{i=1}^{S} p_i \log(p_i)}{\log(\sum_{i=1}^{S} I(x_i > 0))}$
- Sheldon evenness = $\frac{exp(-\sum_{i=1}^{S} p_i \log(p_i))}{\sum_{i=1}^{S} I(x_i > 0)}$
- Beta diversity will be computed using Bray-Curtis dissimilarity.

The difference between groups will be assessed using the Wilcoxon rank-sum test.

```{r}

bc_mat <- tmp |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  as.matrix()

```


```{r}

bc_per_sample <- vegan::vegdist(bc_mat, method = "bray") |>
  as.matrix() |>
  rowMeans() |>
  as_tibble(rownames = "sample_id") |>
  dplyr::rename(bc = value)

```


```{r}

metrics <- tmp |>
  group_by(sample_id) |>
  summarise(
    richness = sum(rel_ab > 0),
    shannon  = vegan::diversity(rel_ab, index = "shannon"),
    pielou   = shannon / log(richness),
    sheldon  = exp(shannon) / richness,
    .groups  = "drop"
  ) |>
  left_join(bc_per_sample, by = "sample_id") |>
  left_join(
    mae[["CDI"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, status),
    by = join_by(sample_id)
  ) |>
  group_by(sample_id, status) |>
  summarise(across(c(richness, shannon, pielou, sheldon, bc), mean), .groups = "drop")

```


```{r}
# richness plot

metrics |> 
  ggplot() +
  aes(x = status, y = richness, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Richness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# shannon diversity plot

metrics |> 
  ggplot() +
  aes(x = status, y = shannon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Shannon diversity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Pielou evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = pielou, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Pielou evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}
# Sheldon evenness plot

metrics |> 
  ggplot() +
  aes(x = status, y = sheldon, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Sheldon evenness", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```


```{r}

# Bray Curtis plot

metrics |> 
  ggplot() +
  aes(x = status, y = bc, fill = status, color = status) +
  geom_violin(color = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.7, size = 1.5) +
  labs(x = "", y = "Bray-Curtis dissimilarity", fill = "Status", color = "Status") +
    stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  )

```