---
title: "Network inference"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| include: false
#| label: setup

library(readr)
library(biomformat) # BiocManager::install("biomformat")
library(stringr)
library(remotes)
library(phyloseq)
library(MicrobeDS) # remotes::install_github("twbattaglia/MicrobeDS")
library(readxl)
library(tibble)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(gt)
library(vegan)
library(quarto)
library(ggpubr)
library(seqtime) # remotes::install_github("hallucigenia-sparsa/seqtime")
library(beemStatic) # remotes::install_github('lch14forever/beem-static')
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
# library(tidySummarizedExperiment) # BiocManager::install("tidySummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

tmp <- fs::dir_map("R", source)
rm(tmp)

theme_set(theme_light())

```

In this file, we will perform network inference on the gut microbiota datasets using three different methods: LIMITS, FlashWeave and BEEM-Static. We will first import the relevant datasets from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI) as well as the healthy cohorts from the Human Microbiome Project (HMPv13 and HMPv35), NIBIOHN and MORINAGA cohorts, and a "diet intervention" cohort as well. We describe first each tool and its implementation in the following sections, and then we will compare the results obtained from each method in a final section.

For each tool, the following metrics will be computed:

* **Number of edges**: The total number of edges in the inferred network.

* **Proportion of positive edges (PEP)**: Defined as the number of positive edges divided by the total number of edges in the inferred network.

* **Richness**: The number of taxa in the microbiota dataset.

* **Shannon diversity**: Defined as $H' = -\sum_{i=1}^S p_i \ln(p_i)$, where $p_i$ is the relative abundance of taxon $i$ and $S$ is the total number of taxa in the dataset.

* **Sheldon evenness**: Defined as $E = \frac{H'}{\ln(S)}$, where $H'$ is the Shannon diversity and $S$ is the total number of taxa in the dataset.

* **Net interactions**: computed as the difference between the sum of positive and negative edge weights, normalized by the total sum of edge weights in the inferred network. This metric quantifies the balance between positive and negative interactions, ranging from -1 (all interactions negative) to +1 (all interactions positive). Mathematically:

$$\rho = \frac{\sum_{i \neq j} (w_{ij}^+ + w_{ij}^-)}{\sum_{i \neq j} (w_{ij}^+ - w_{ij}^-)}$$

where $w_{ij}^+$ and $w_{ij}^-$ denote the positive and negative edge weights between taxa i and j respectively.For directed inference tools (BEEM-Static, LIMITS), both directions $w_{ij}$ and $w_{ji}$ are included. For undirected tools (FlashWeave), each pair is counted once. The ratio remains comparable across tools as any double-counting cancels in the ratio.

# Importing the MAE object

```{r}

mae_file <- 
  fs::dir_ls(str_c(data_dir(), "/02-processed-data"), regexp = "MAE_processed_.*\\.rds$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)


cat(str_c("Loading MultiAssayExperiment from: 00_data/", mae_file |> str_remove(data_dir()), "\n"))

mae <- readRDS(mae_file)

rm(mae_file)

mae

```

# Network inference for main cohorts

## Network inference for IBD cohort (Lloyd-Price et al. 2019)

```{r}

tmp <-
  mae[["IBD"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["IBD"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, pid, visit_num, diagnosis),
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(diagnosis == "nonIBD")
tmp_diseased <- tmp |> filter(diagnosis == "UC" | diagnosis == "CD")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In this section, we will apply FlashWeave to the different groups (healthy and diseased) separately.

FlashWeave is a network inference tool developed by the Von Mering research group (Tackmann et al., 2019). It operates within a Probabilistic Graphical Model framework, enabling the integration of metadata and the prediction of direct microbial interactions from heterogeneous microbial abundance datasets.

We first begin by loading the Julia package FlashWeave using the `JuliaCall` package in R. This package enables us to run Julia code from R.

```{r}
#| include: true
#| eval: true

library(JuliaCall)
julia_setup(JULIA_HOME = "C:/Users/maria/.julia/juliaup/julia-1.12.5+0.x64.w64.mingw32/bin")
julia_eval('using FlashWeave')
julia_eval('import FlashWeave: learn_network, save_network, load_data')
julia_eval('println(methods(learn_network))')

```

As per the original article, FlashWeave was run with its default parameters, and `heterogeneous = FALSE` was chosen, in accordance with the original article. We run FlashWeave on bootstrap iterations with the function `run_network_bootstrap()` that we defined in the R/ directory. This function takes as input a dataset in matrix form, a network inference function (in this case, `run_flashweave()`, which is also defined in the R/ directory), and the parameters for the bootstrap iterations. The function returns a list of results for each bootstrap iteration, which can be used to compute the statistics of interest (PEP, Sheldon evenness, Shannon diversity, richness, and net interactions), summarized with the function `get_stats_df()`. The parameters for the bootstrap iterations are as follows: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`, and no prevalence threshold. Additionally, the parameters from `get_stats_df()` are set to `symmetric = TRUE` since FlashWeave returns undirected edges.

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

```

We save the results in an RDS file for later use.

```{r}
#| eval: false

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_IBD_20260419.rds"))

beepr::beep(1)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_IBD_20260419.rds"))

beepr::beep(1)

```

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_IBD_20260419.rds"))

res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_IBD_20260419.rds"))

```

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```

```{r}
#| fig-height: 10
#| fig-width: 12
# PEP plot

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group",
       subtitle = "Comparison of PEP between healthy and diseased \nsamples in the IBD cohort") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group",
       subtitle = "Comparison of evenness between healthy and \ndiseased samples in the IBD cohorts") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group",
       subtitle = "Comparison of Shannon diversity between healthy and \ndiseased samples in the IBD cohort") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group",
       subtitle = "Comparison of richness between healthy and \ndiseased samples in the IBD cohort") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + patchwork::plot_annotation(title = "FlashWeave results for the IBD cohort (Lloyd-Price et al., 2019)", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.015,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group",
       title = "FlashWeave results for the IBD cohort (Lloyd-Price et al., 2019)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```



```{r}
#| fig-height: 6
#| fig-width: 12


df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +

  df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
patchwork::plot_layout(nrow = 1) + patchwork::plot_annotation(title = "FlashWeave results for the IBD cohort (Lloyd-Price et al., 2019)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```



```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.015,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group", 
       title = "FlashWeave results for the IBD cohort (Lloyd-Price et al., 2019)", subtitle = "Correlation between Sheldon evenness and net interactions in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```

### BEEM-Static

```{r}
# import file from cluster

res_healthy <- readRDS(file = str_c(result_dir(), "/00-raw-beem-static/beem_IBD_healthy_20260414.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/00-raw-beem-static/beem_IBD_diseased_20260414.rds"))

```

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = FALSE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = FALSE)

```

```{r}

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```

```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(ncol = 2) + patchwork::plot_annotation(title = "BEEM-Static results for the IBD cohort (Lloyd-Price et al., 2019)", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```

## Network inference for IBS cohort (Mars et al. 2020)

```{r}

tmp <-
  mae[["IBS"]] |> 
  assay() |> 
  t() |> 
  as.data.frame() |> 
  rownames_to_column("uid") |> 
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["IBS"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, pid, time, cohort), 
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(cohort == "H")
tmp_diseased <- tmp |> filter(cohort == "C" | cohort == "D")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data.

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

beepr::beep(2)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_IBS_20260419.rds"))
saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_IBS_20260419.rds"))

beepr::beep(2)

```

```{r}

# res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_IBS_20260419.rds"))
# res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_IBS_20260419.rds"))

```

```{r}

# df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
# df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)
# 
# df_res <-
#   bind_rows(
#     df_res_healthy |> mutate(group = "Healthy"),
#     df_res_diseased |> mutate(group = "Diseased")
#   )

```

```{r}
#| fig-height: 10
#| fig-width: 12
# PEP plot

# df_res |>
#   ggplot() +
#   aes(x = group, y = pep, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group",
#        subtitle = "Comparison of PEP between healthy and diseased \nsamples in the IBS cohort") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
#   
#   df_res |>
#   ggplot() +
#   aes(x = group, y = sheldon, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group",
#        subtitle = "Comparison of evenness between healthy and \ndiseased samples in the IBS cohorts") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
#   
# # shannon diversity
# df_res |>
#   ggplot() +
#   aes(x = group, y = shannon, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group",
#        subtitle = "Comparison of Shannon diversity between healthy and \ndiseased samples in the IBS cohort") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
#   
# # richness
# df_res |>
#   ggplot() +
#   aes(x = group, y = shannon, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Richness", color = "Group", fill = "Group",
#        subtitle = "Comparison of richness between healthy and \ndiseased samples in the IBS cohort") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
#   
#   patchwork::plot_layout(nrow = 2) + patchwork::plot_annotation(title = "FlashWeave results for the IBS cohort (Mars et al. 2020)", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```

> ...

We will now check if there is any correlation between Sheldon evenness and PEP in the plot below.

```{r}

# df_res |>
#   ggplot() +
#   aes(x = sheldon, y = pep, color = group, fill = group) +
#   geom_point(alpha = 0.6, size = 1.5) +
#   geom_smooth(method = "lm", se = TRUE) +
#   geom_smooth(aes(x = sheldon, y = pep), method = "lm",
#               se = TRUE, inherit.aes = FALSE,
#               color = "black", linetype = "dashed") +
#   ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
#                    method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
#                    size = 3, show.legend = FALSE) +
#   ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
#                    inherit.aes = FALSE, color = "black",
#                    label.x.npc = "left", label.y.npc = 0.015,
#                    size = 3) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group",
#        title = "FlashWeave results for the IBS cohort (Mars et al. 2020)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```

> ...

We also check if the results for net interactions (rho) are consistent with the PEP results, and if there is any significant difference between healthy and diseased samples in terms of net interactions.

```{r}
#| fig-height: 6
#| fig-width: 12


# df_res |>
#   ggplot() +
#   aes(x = group, y = rho, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
# 
#   df_res |>
#   ggplot() +
#   aes(x = group, y = pep, color = group, fill = group) +
#   geom_violin(alpha = 0.4, color = NA) +
#   geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
#   stat_compare_means(
#     comparisons = list(c("Healthy", "Diseased")),
#     method = "wilcox.test",
#     label = "p.signif",
#     tip.length = 0.01,
#     bracket.size = 0.8,
#     fontface = "bold"
#   ) +
#   
# patchwork::plot_layout(nrow = 1) + patchwork::plot_annotation(title = "FlashWeave results for the IBS cohort (Mars et al. 2020)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```

> ...

We will check correlation between net interaction and Sheldon evenness in the plot below.

```{r}

# df_res |>
#   ggplot() +
#   aes(x = sheldon, y = rho, color = group, fill = group) +
#   geom_point(alpha = 0.6, size = 1) +
#   geom_smooth(method = "lm", se = TRUE) +
#   geom_smooth(aes(x = sheldon, y = rho), method = "lm",
#               se = TRUE, inherit.aes = FALSE,
#               color = "black", linetype = "dashed") +
#   ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
#                    method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
#                    size = 3, show.legend = FALSE) +
#   ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
#                    inherit.aes = FALSE, color = "black",
#                    label.x.npc = "left", label.y.npc = 0.015,
#                    size = 3) +
#   scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
#   facet_wrap(~ fraction) +
#   labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group", 
#        title = "FlashWeave results for the IBS cohort (Mars et al. 2020)", subtitle = "Correlation between Sheldon evenness and net interactions in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```

### BEEM-Static

```{r}

```

## Network inference for CDI cohort (Ferretti et al., 2023)

```{r}

tmp <-
  mae[["CDI"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CDI"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, status, subject_disease_status),
    by = join_by(uid)
  ) |>
  filter(subject_disease_status == "CDI" | subject_disease_status == "CTR")


```

```{r}

tmp_healthy <- tmp |> filter(subject_disease_status == "CTR")
tmp_diseased <- tmp |> filter(subject_disease_status == "CDI")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data. 

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

beepr::beep(2)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_CDI_20260419.rds"))
saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_CDI_20260419.rds"))

beepr::beep(2)

```

We load the results in the chunk code below.

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_CDI_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_CDI_20260419.rds"))

```

The results are obtained in the chunk code below.

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```


```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +  
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +  
  
  patchwork::plot_layout(nrow = 2) + patchwork::plot_annotation(title = "FlashWeave results for the CDI cohort (Ferretti et al., 2023)", subtitle = "Comparison of PEP, evenness, Shannon diversity and richness between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}
# we rank total relative abundances for each taxa and check the most abundant ones

tmp_healthy |>
  group_by(taxon) |>
  summarise(total_rel_ab = sum(rel_ab)) |>
  mutate(proportion = total_rel_ab / sum(total_rel_ab)) |>
  arrange(desc(proportion)) |>
  select(taxon, proportion) |>
  slice_head(n = 5) |>
  gt(caption = "Top 5 most abundant taxa in healthy CDI samples (proportion of total relative abundance)") |>
  fmt_percent(columns = "proportion", decimals = 2)

tmp_diseased |>
  group_by(taxon) |>
  summarise(total_rel_ab = sum(rel_ab)) |>
  mutate(proportion = total_rel_ab / sum(total_rel_ab)) |>
  arrange(desc(proportion)) |>
  select(taxon, proportion) |>
  slice_head(n = 5) |>
  gt(caption = "Top 5 most abundant taxa in diseased CDI samples (proportion of total relative abundance)") |>
  fmt_percent(columns = "proportion", decimals = 2)


```


We will now check if there is any correlation between Sheldon evenness and PEP in the plot below.

```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.010,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the CDI cohort (Ferretti et al., 2023)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```

We also compute the net interactions (rho) in the healthy and diseased samples and compare it with PEP in the plot below.

```{r}
#| fig-height: 6
#| fig-width: 12


df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the CDI cohort (Ferretti et al., 2023)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```

We check correlation between net interaction and Sheldon evenness in the plot below.

```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.12,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.01,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group", title = "FlashWeave results for the CDI cohort (Ferretti et al., 2023)", subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Network inference for CRC cohort (Yachida et al., 2019)

```{r}

tmp <-
  mae[["CRC"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CRC"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, cohort),
    by = join_by(uid)
  )

```


```{r}

tmp_healthy <- tmp |> filter(cohort == "H")
tmp_diseased <- tmp |> filter(cohort == "CRC")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data. 

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_CRC_20260419.rds"))

beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_CRC_20260419.rds"))

beepr::beep(2)

```

We load the results in the chunk code below.

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_CRC_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_CRC_20260419.rds"))

```

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```

We plot the results below.

```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + plot_annotation(title = "FlashWeave results for the CRC cohort (Yachida et al., 2019)", subtitle = "Comparison between healthy and diseased samples for PEP, Sheldon evenness, Shannon diversity and richness", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the CRC cohort (Yachida et al., 2019)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```


```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the CRC cohort (Yachida et al., 2019)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group",
       title = "FlashWeave results for the CRC cohort (Yachida et al., 2019)",
       subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples",
       caption = "Correlation is computed with Pearson's correlation test")

```

# Network inference for supplementary cohorts

## Network inference for CRC cohort (Zeller et al., 2014)

```{r}

tmp <-
  mae[["CRC_ZELLER"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CRC_ZELLER"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, disease),
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(disease == "healthy")
tmp_diseased <- tmp |> filter(disease == "CRC")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data. 

```{r}
#| eval: false
res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_CRC_zeller_20260419.rds"))

beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_CRC_zeller_20260419.rds"))

beepr::beep(2)

```

We load the results in the chunk code below.

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_CRC_zeller_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_CRC_zeller_20260419.rds"))

```


```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```


```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + plot_annotation(title = "FlashWeave results for the CRC cohort (Yachida et al., 2019)", subtitle = "Comparison between healthy and diseased samples for PEP, Sheldon evenness, Shannon diversity and richness", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the CRC cohort (Zeller et al., 2014)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```


```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the CRC cohort (Zeller et al., 2014)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group",
       title = "FlashWeave results for the CRC cohort (Zeller et al., 2014)",
       subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples",
       caption = "Correlation is computed with Pearson's correlation test")

```


## Network inference for IBD cohort (Franzosa et al., 2019)

```{r}

tmp <-
  mae[["IBD_FRANZOSA"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "rel_ab") |>
  # group_by(uid) |>
  # mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["IBD_FRANZOSA"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, diagnosis),
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(diagnosis == "Control")
tmp_diseased <- tmp |> filter(diagnosis == "CD" | diagnosis == "UC")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data. 

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_IBD_franzosa_20260419.rds"))

beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_IBD_franzosa_20260419.rds"))

beepr::beep(2)

```

We load the results in the chunk code below.

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_IBD_franzosa_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_IBD_franzosa_20260419.rds"))

```

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```


```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + plot_annotation(title = "FlashWeave results for the IBD cohort (Franzosa et al., 2019)", subtitle = "Comparison between healthy and diseased samples for PEP, Sheldon evenness, Shannon diversity and richness", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the IBD cohort (Franzosa et al., 2019)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```


```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the IBD cohort (Franzosa et al., 2019)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```

```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group",
       title = "FlashWeave results for the IBD cohort (Franzosa et al., 2019)",
       subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples",
       caption = "Correlation is computed with Pearson's correlation test")

```


## Network inference for CRC cohort (Wirbel et al., 2018)

```{r}

tmp <-
  mae[["CRC_WIRBEL"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CRC_WIRBEL"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, disease),
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(disease == "healthy")
tmp_diseased <- tmp |> filter(disease == "CRC")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_CRC_wirbel_20260419.rds"))
beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_CRC_wirbel_20260419.rds"))
beepr::beep(2)

```

```{r}
# load results

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_CRC_wirbel_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_CRC_wirbel_20260419.rds"))

```

```{r}
df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )
```

```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + plot_annotation(title = "FlashWeave results for the CRC cohort (Wirbel et al., 2018)", subtitle = "Comparison between healthy and diseased samples for PEP, Sheldon evenness, Shannon diversity and richness", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the CRC cohort (Wirbel et al., 2018)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```


```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the CRC cohort (Wirbel et al., 2018)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group",
       title = "FlashWeave results for the CRC cohort (Wirbel et al., 2018)",
       subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples",
       caption = "Correlation is computed with Pearson's correlation test")

```

## Network inference for CRC cohort (Vogtmann et al., 2016)

```{r}

tmp <-
  mae[["CRC_VOGTMANN"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "count") |>
  group_by(uid) |>
  mutate(rel_ab = count/sum(count)) |>
  left_join(
    mae[["CRC_VOGTMANN"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, disease),
    by = join_by(uid)
  )

```


```{r}

tmp_healthy <- tmp |> filter(disease == "healthy")
tmp_diseased <- tmp |> filter(disease == "CRC")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

### FlashWeave

```{r}
#| eval: false

res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_CRC_vogtmann_20260419.rds"))
beepr::beep(2)

res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0)

saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_CRC_vogtmann_20260419.rds"))
beepr::beep(2)

```


```{r}
# load results

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_CRC_vogtmann_20260419.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_CRC_vogtmann_20260419.rds"))

```


```{r}
df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )
```


```{r}
#| fig-height: 10
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
df_res |>
  ggplot() +
  aes(x = group, y = sheldon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Sheldon evenness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# shannon diversity
df_res |>
  ggplot() +
  aes(x = group, y = shannon, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Shannon diversity", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
# richness
df_res |>
  ggplot() +
  aes(x = group, y = richness, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Richness", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 2) + plot_annotation(title = "FlashWeave results for the CRC cohort (Vogtmann et al., 2016)", subtitle = "Comparison between healthy and diseased samples for PEP, Sheldon evenness, Shannon diversity and richness", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = pep, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = pep), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = pep), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group", title = "FlashWeave results for the CRC cohort (Vogtmann et al., 2016)", subtitle = "Correlation between Sheldon evenness and PEP in healthy and diseased samples", caption = "Correlation is computed with Pearson's correlation test")

```


```{r}
#| fig-height: 6
#| fig-width: 12

df_res |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
df_res |>
  ggplot() +
  aes(x = group, y = pep, color = group, fill = group) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "", y = "Positive edge percentage (PEP)", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold"
  ) +
  
  patchwork::plot_layout(nrow = 1) + plot_annotation(title = "FlashWeave results for the CRC cohort (Vogtmann et al., 2016)", subtitle = "Comparison between PEP and net interaction results between healthy and diseased samples", caption = "*: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001; ns: not significant (Wilcoxon test)")

```


```{r}

df_res |>
  ggplot() +
  aes(x = sheldon, y = rho, color = group, fill = group) +
  geom_point(alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_smooth(aes(x = sheldon, y = rho), method = "lm",
              se = TRUE, inherit.aes = FALSE,
              color = "black", linetype = "dashed") +
  ggpubr::stat_cor(aes(label = after_stat(paste(r.label, p.label, sep = "~`,`~"))),
                   method = "pearson", label.x.npc = "left", label.y.npc = 0.10,
                   size = 3, show.legend = FALSE) +
  ggpubr::stat_cor(aes(x = sheldon, y = rho), method = "pearson",
                   inherit.aes = FALSE, color = "black",
                   label.x.npc = "left", label.y.npc = 0.001,
                   size = 3) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  facet_wrap(~ fraction) +
  labs(x = "Sheldon evenness", y = "Net interactions (rho)", color = "Group", fill = "Group",
       title = "FlashWeave results for the CRC cohort (Vogtmann et al., 2016)",
       subtitle = "Correlation between Sheldon evenness and net interaction in healthy and diseased samples",
       caption = "Correlation is computed with Pearson's correlation test")

```
