---
title: "Network inference"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| include: false
#| label: setup

library(readr)
library(biomformat) # BiocManager::install("biomformat")
library(stringr)
library(remotes)
library(phyloseq)
library(MicrobeDS) # remotes::install_github("twbattaglia/MicrobeDS")
library(readxl)
library(tibble)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(gt)
library(vegan)
library(quarto)
library(ggpubr)
library(seqtime) # remotes::install_github("hallucigenia-sparsa/seqtime")
library(beemStatic) # remotes::install_github('lch14forever/beem-static')
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
# library(tidySummarizedExperiment) # BiocManager::install("tidySummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

tmp <- fs::dir_map("R", source)
rm(tmp)

theme_set(theme_light())

```

In this file, we will perform network inference on the gut microbiota datasets using three different methods: LIMITS, FlashWeave and BEEM-Static. We will first import the relevant datasets from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI) as well as the healthy cohorts from the Human Microbiome Project (HMPv13 and HMPv35), NIBIOHN and MORINAGA cohorts, and a "diet intervention" cohort as well. We describe first each tool and its implementation in the following sections, and then we will compare the results obtained from each method in a final section.

For each tool, the following metrics will be computed:

* **Number of edges**: The total number of edges in the inferred network.

* **Proportion of positive edges (PEP)**: Defined as the number of positive edges divided by the total number of edges in the inferred network.

* **Richness**: The number of taxa in the microbiota dataset.

* **Shannon diversity**: Defined as $H' = -\sum_{i=1}^S p_i \ln(p_i)$, where $p_i$ is the relative abundance of taxon $i$ and $S$ is the total number of taxa in the dataset.

* **Sheldon evenness**: Defined as $E = \frac{H'}{\ln(S)}$, where $H'$ is the Shannon diversity and $S$ is the total number of taxa in the dataset.

* **Net interactions**: computed as the difference between the sum of positive and negative edge weights, normalized by the total sum of edge weights in the inferred network. This metric quantifies the balance between positive and negative interactions, ranging from -1 (all interactions negative) to +1 (all interactions positive). Mathematically:

$$\rho = \frac{\sum_{i \neq j} (w_{ij}^+ + w_{ij}^-)}{\sum_{i \neq j} (w_{ij}^+ - w_{ij}^-)}$$

where $w_{ij}^+$ and $w_{ij}^-$ denote the positive and negative edge weights between taxa i and j respectively.For directed inference tools (BEEM-Static, LIMITS), both directions $w_{ij}$ and $w_{ji}$ are included. For undirected tools (FlashWeave), each pair is counted once. The ratio remains comparable across tools as any double-counting cancels in the ratio.


# Importing the MAE object

```{r}

mae_file <- 
  fs::dir_ls(str_c(data_dir(), "/02-processed-data"), regexp = "MAE_processed_.*\\.rds$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)


cat(str_c("Loading MultiAssayExperiment from: 00_data/", mae_file |> str_remove(data_dir()), "\n"))

mae <- readRDS(mae_file)

rm(mae_file)

mae

```

# Network inference with LIMITS

```{r 20-network-inference, child = '21-LIMITS.qmd'}
#| eval: false
```

# Network inference with FlashWeave

```{r 20-network-inference, child = '22-FlashWeave.qmd'}
#| eval: true
```

# Network inference with BEEM-Static

```{r 20-network-inference, child = '23-BEEM_Static.qmd'}
#| eval: false
```