---
title: "Network inference with LIMITS"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

In this section, we will apply LIMITS to the IBD cohort only (since it is the only cohort that has enough time points to apply the method).

## IBD cohort

> A sample (starting with H) has a taxon with relative abundance == 1.

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["IBD"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["IBD"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```

### Prevalence filtering

We will then filter out low-abundance taxa, by keeping only taxa that are present in at least 5% of the samples, and the remaining is grouped in a category named "Other". This is done in the chunk code below.

```{r}

tmp_healthy <- tmp |> filter(diagnosis == "nonIBD")
tmp_diseased <- tmp |> filter(diagnosis == "UC" | diagnosis == "CD")

```


```{r}
# for healthy samples

tmp_filtered_healthy <- 
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.05) |>
  left_join(
    mae[["IBD"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```


```{r}
# for diseased samples

tmp_filtered_diseased <- 
  filter_prevalence(tmp_diseased, prevalence_threshold = 0.05) |>
  left_join(
    mae[["IBD"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```

### Network inference with LIMITS (test run)

```{r}

# for healthy samples

healthy_df <-
  tmp_filtered_healthy |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

```

```{r}

# for diseased samples

diseased_df <-
  tmp_filtered_diseased |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

```

```{r}

res <- run_limits_bootstrap(healthy_df, n_bootstrap = 1, subsample_pcts = 0.4, max_attempts = 5, bagging.iter = 5)

```

```{r}

df_res <- get_limits_results(res)

```

