---
title: "Network inference with BEEM-Static"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

In this section, we will apply BEEM-Static to the differents cohorts from the main analysis, and additionnaly compute beta diversity, evenness, and richness measures.

## IBD cohort

> A sample (starting with H) has a taxon with relative abundance == 1.

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["IBD"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["IBD"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```

### Prevalence filtering

We will then filter out low-abundance taxa, by keeping only taxa that are present in at least 20% of the samples, and the remaining is grouped in a category named "Other". This is done in the chunk code below.

::: {.callout-note}
The choice of a prevalence threshold of 20% is somewhat arbitrary, but it is necessary in this case since BEEM-Static does not handle datasets with lots of zeros well.
:::

```{r}

tmp_healthy <- tmp |> filter(diagnosis == "nonIBD")
tmp_diseased <- tmp |> filter(diagnosis == "UC" | diagnosis == "CD")

```


```{r}
# for healthy samples

tmp_filtered_healthy <- 
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.2) |>
  left_join(
    mae[["IBD"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```


```{r}
# for diseased samples

tmp_filtered_diseased <- 
  filter_prevalence(tmp_diseased, prevalence_threshold = 0.2) |>
  left_join(
    mae[["IBD"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, participant_id, visit_num, diagnosis),
    by = join_by(sample_id)
  )

```

### Network inference (test run)

BEEM Static expect as input a matrix of counts, with rows as taxa and columns as samples. We will thus transform the filtered dataset into the expected format, and then apply BEEM-Static to infer the microbial association network.

We will first transform the filtered datasets (healthy and diseased) into the expected format, and then apply BEEM-Static to infer the microbial association network. This is done in the chunk code below.

```{r}
# for healthy samples

healthy_df <-
  tmp_filtered_healthy |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  healthy_df, file.path(result_dir(), "00-raw-beem-static", str_c("IBD_healthy_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```


```{r}
# for diseased samples

diseased_df <-
  tmp_filtered_diseased |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  diseased_df, file.path(result_dir(), "00-raw-beem-static", str_c("IBD_diseased_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```

We apply the algorithm on the healthy and diseased datasets separately, to infer two separate networks.

```{r}
#| eval: false

res_healthy <- beemStatic::func.EM(healthy_df, max.iter = 2)

```

The results for the healthy dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_healthy)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```


```{r}
#| eval: false

res_diseased <- beemStatic::func.EM(diseased_df)

```

The results for diseased dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_diseased)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```

## IBS cohort

We first import the IBS dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["IBS"]] |> 
  assay() |> 
  t() |> 
  as.data.frame() |> 
  rownames_to_column("sample_id") |> 
  pivot_longer(-sample_id, names_to = "taxon", values_to = "counts") |>
  group_by(sample_id) |>
  mutate(rel_ab = counts / sum(counts)) |>
  ungroup() |>
  left_join(
    mae[["IBS"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, study_id, timepoint, cohort), 
    by = join_by(sample_id)
  )

```


### Prevalence filtering

We will then filter out low-abundance taxa, by keeping only taxa that are present in at least 50% of the samples, and the remaining is grouped in a category named "Other". This is done in the chunk code below.

::: {.callout-note}
The choice of a prevalence threshold of 50% is somewhat arbitrary, but it is necessary in this case since BEEM-Static does not handle datasets with lots of zeros well.
:::

```{r}

tmp_healthy <- tmp |> filter(cohort == "H")
tmp_diseased <- tmp |> filter(cohort == "IBS")

```


```{r}
# for healthy samples

tmp_filtered_healthy <-
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.5) |>
  left_join(
    mae[["IBS"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, study_id, timepoint, cohort),
    by = join_by(sample_id)
  )
    
```


```{r}
# for diseased samples

tmp_filtered_diseased <-
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.5) |>
  left_join(
    mae[["IBS"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, study_id, timepoint, cohort),
    by = join_by(sample_id)
  )

```

### Network inference (test run)

BEEM Static expect as input a matrix of relative abundances, with rows as taxa and columns as samples. We will thus transform the filtered dataset into the expected format, and then apply BEEM-Static to infer the microbial association network.

We will first transform the filtered datasets (healthy and diseased) into the expected format, and then apply BEEM-Static to infer the microbial association network. This is done in the chunk code below.

```{r}
# for healthy samples

healthy_df <-
  tmp_filtered_healthy |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  healthy_df, file.path(result_dir(), "00-raw-beem-static", str_c("IBS_healthy_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```


```{r}
# for diseased samples

diseased_df <-
  tmp_filtered_diseased |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  diseased_df, file.path(result_dir(), "00-raw-beem-static", str_c("IBS_diseased_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```

We apply the algorithm on the healthy and diseased datasets separately, to infer two separate networks.

```{r}
#| eval: false

res_healthy <- beemStatic::func.EM(healthy_df)

```

The results for the healthy dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_healthy)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```


```{r}
#| eval: false

res_diseased <- beemStatic::func.EM(diseased_df)

```

The results for the diseased dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_diseased)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```

## CRC cohort (Ferretti et al., 2022)

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["CRC"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["CRC"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, group, cohort),
    by = join_by(sample_id)
  )

```

### Prevalence filtering

We will then filter out low-abundance taxa, by keeping only taxa that are present in at least 20% of the samples, and the remaining is grouped in a category named "Other". This is done in the chunk code below.

::: {.callout-note}
The choice of a prevalence threshold of 20% is somewhat arbitrary, but it is necessary in this case since BEEM-Static does not handle datasets with lots of zeros well.
:::

```{r}

tmp_healthy <- tmp |> filter(cohort == "H")
tmp_diseased <- tmp |> filter(cohort == "CRC")

```


```{r}
# for healthy samples

tmp_filtered_healthy <- 
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.2) |>
  left_join(
    mae[["CRC"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, group, cohort),
    by = join_by(sample_id)
  )

```


```{r}
# for diseased samples

tmp_filtered_diseased <-
  filter_prevalence(tmp_diseased, prevalence_threshold = 0.2) |>
  left_join(
    mae[["CRC"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, group, cohort),
    by = join_by(sample_id)
  )

```

### Network inference (test run)

BEEM Static expect as input a matrix of relative abundances, with rows as taxa and columns as samples. We will thus transform the filtered dataset into the expected format, and then apply BEEM-Static to infer the microbial association network.

We will first transform the filtered datasets (healthy and diseased) into the expected format, and then apply BEEM-Static to infer the microbial association network. This is done in the chunk code below.

```{r}
# for healthy samples

healthy_df <-
  tmp_filtered_healthy |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  healthy_df, file.path(result_dir(), "00-raw-beem-static", str_c("CRC_healthy_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```


```{r}
# for diseased samples

diseased_df <-
  tmp_filtered_diseased |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  diseased_df, file.path(result_dir(), "00-raw-beem-static", str_c("CRC_diseased_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```

We apply the algorithm on the healthy and diseased datasets separately, to infer two separate networks.

```{r}
#| eval: false

res_healthy <- beemStatic::func.EM(healthy_df)

```

The results for the healthy dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_healthy)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```


```{r}
#| eval: false

res_diseased <- beemStatic::func.EM(diseased_df)

```

The results for the diseased dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_diseased)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```

## CDI cohort

> Skip for this one and justify why. 90% sparsity!

We first import the IBD dataset from the MAE object in the chunk code below.

```{r}

tmp <-
  mae[["CDI"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "counts") |>
  group_by(sample_id) |>
  mutate(rel_ab = counts / sum(counts)) |>
  ungroup() |>
  left_join(
    mae[["CDI"]] |> colData() |> as_tibble(rownames = "sample_id") |> select(sample_id, status, subject_disease_status),
    by = join_by(sample_id)
  )

```

### Prevalence filtering

We will then filter out low-abundance taxa, by keeping only taxa that are present in at least 50% of the samples, and the remaining is grouped in a category named "Other". This is done in the chunk code below.

::: {.callout-note}
The choice of a prevalence threshold of 50% is somewhat arbitrary, but it is necessary in this case since BEEM-Static does not handle datasets with lots of zeros well.
:::

```{r}

tmp_healthy <- tmp |> filter(status == "Healthy")
tmp_diseased <- tmp |> filter(status == "Diseased")

```


```{r}
# for healthy samples

tmp_filtered_healthy <-
  filter_prevalence(tmp_healthy, prevalence_threshold = 0.6) |>
  left_join(
    mae[["CDI"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, status),
    by = join_by(sample_id)
  )

```


```{r}
# for diseased samples

tmp_filtered_diseased <-
  filter_prevalence(tmp_diseased, prevalence_threshold = 0.6) |>
  left_join(
    mae[["CDI"]] |> colData() |> as.data.frame() |> as_tibble(rownames = "sample_id") |> 
      select(sample_id, status),
    by = join_by(sample_id)
  )

```

### Network inference (test run)

BEEM Static expect as input a matrix of relative abundances, with rows as taxa and columns as samples. We will thus transform the filtered dataset into the expected format, and then apply BEEM-Static to infer the microbial association network.

We will first transform the filtered datasets (healthy and diseased) into the expected format, and then apply BEEM-Static to infer the microbial association network. This is done in the chunk code below.

```{r}
# for healthy samples

healthy_df <-
  tmp_filtered_healthy |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  healthy_df, file.path(result_dir(), "00-raw-beem-static", str_c("CDI_healthy_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```


```{r}
# for diseased samples

diseased_df <-
  tmp_filtered_diseased |>
  select(sample_id, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("sample_id") |>
  t()

saveRDS(
  diseased_df, file.path(result_dir(), "00-raw-beem-static", str_c("CDI_diseased_BEEM_", today() |> str_remove_all("-"), ".rds"))
  )

```

We apply the algorithm on the healthy and diseased datasets separately, to infer two separate networks.

```{r}
#| eval: false

res_healthy <- beemStatic::func.EM(healthy_df)

```

The results for the healthy dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_healthy)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```


```{r}
#| eval: false

res_diseased <- beemStatic::func.EM(diseased_df)

```

The results for the diseased dataset are displayed below.

```{r}
#| eval: false

est <- beem2param(res_diseased)
pep_inferred <- seqtime::getPep(est$b.est)

# non-zero entries in the upper and lower triangle (skipping diagonal) 
values <- est$b.est[upper.tri(est$b.est)]
values <- c(values,est$b.est[lower.tri(est$b.est)])

edge_num <- (length(which(values != 0))) 

cat("Number of edges in the inferred network:", edge_num, "\n")
cat("Proportion of positive edges in the inferred network:", pep_inferred, "\n")

```