---
title: "Comments on 'Imbalance in gut microbial interactionsas a marker of health and disease' from Corral López et al. (2026)"
bibliography: references.bib
date: today 
format:
  html:
    page-layout: full
    code-fold: false
    code-tools: false
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  echo: false
  cache: false 
  warning: false
project:
  execute-dir: project
  type: website
editor: source
editor_options: 
  chunk_output_type: inline
---

```{r}
#| include: false
#| label: setup

library(readr)
library(biomformat) # BiocManager::install("biomformat")
library(stringr)
library(remotes)
library(phyloseq)
library(MicrobeDS) # remotes::install_github("twbattaglia/MicrobeDS")
library(readxl)
library(tibble)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(gt)
library(vegan)
library(quarto)
library(ggpubr)
library(seqtime) # remotes::install_github("hallucigenia-sparsa/seqtime")
library(beemStatic) # remotes::install_github('lch14forever/beem-static')
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

tmp <- fs::dir_map("R", source)
rm(tmp)

theme_set(theme_light())

```

```{r}
#| include: false

reset <- FALSE

```


## Introduction

The human gut microbiome is a complex ecological system in which microbial interactions are thought to play a central role in maintaining host health and contributing to disease. In this context, @CorralLopez2026 proposed that an imbalance in microbial interactions, quantified through an ecological index (ENBI), constitutes a universal signature of disease across multiple gut-related conditions, which includes inflammatory bowel disease (IBD), irritable bowel syndrome (IBS), colorectal cancer (CRC), and *Clostridioides difficile* infection (CDI). Specifically, they reported that diseased microbiomes are characterized by a predominance of positive interactions, whereas healthy microbiomes exhibit more negative interactions, suggesting that ENBI could serve as a universal biomarker of disease. Such a claim, if robust, would have important implications for microbiome-based diagnostics and our understanding of microbial ecology in disease. However, as this will be developed, we believe the generalizability of this pattern across datasets and conditions remains to be established.

In this study, we reassessed microbial interaction patterns by quantifying net interactions, defined as the balance between positive and negative associations inferred from microbial networks, as used in the ENBI framework, across multiple independent datasets. Using different network inference methods and resampling strategies, we examined whether the reported interaction imbalance consistently distinguishes healthy and diseased microbiomes across conditions.

## Results and discussion

We analyzed multiple gut microbiome datasets from inflammatory bowel disease (IBD), *Clostridioides difficile* infection (CDI), and colorectal cancer (CRC), including both primary (used by @CorralLopez2026) and independent validation cohorts. Sample sizes and features for each dataset are summarized in @tbl-summary-main and @tbl-summary-supp. All datasets were processed in `00-data-preprocessing.qmd` (see GitHub repository), and microbial interaction networks were inferred using both FlashWeave and BEEM-Static, combined with resampling strategies for bootstrapping.

```{r}

# import latest MAE

mae_file <- 
  fs::dir_ls(str_c(data_dir(), "/02-processed-data"), regexp = "MAE_processed_.*\\.rds$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_file)

rm(mae_file)

```

```{r}
#| include: false

group_var <- c(IBD = "diagnosis", CDI = "group", CRC = "cohort", CRC_VOGTMANN = "disease", CRC_ZELLER = "disease", CRC_WIRBEL = "disease", IBD_FRANZOSA = "diagnosis")

tbl <- map_dfr(names(group_var), function(ds) {
  
  mat <- mae[[ds]] |> assay()
  meta <- colData(mae[[ds]]) |> as.data.frame()
  
  gvar <- group_var[[ds]]
  
  tibble(
    dataset = ds,
    group = meta[[gvar]]
  ) |>
    group_by(dataset, group) |>
    summarise(
      n_samples = n(),
      n_taxa = nrow(mat),
      .groups = "drop"
    )
})

tbl_clean <- tbl |>
  filter(group != "adenoma") |>
  mutate(
    group = case_when(
      group %in% c("CD", "UC") ~ "IBD",
      group %in% c("nonIBD", "Control", "control", "healthy", "H") ~ "healthy",
      TRUE ~ group
    )
  ) |>
  group_by(dataset, group) |>
  summarise(
    n_samples = sum(n_samples),
    n_taxa = dplyr::first(n_taxa),
    .groups = "drop"
  )

tbl_clean <- tbl_clean |>
  mutate(dataset = recode(dataset,
    IBD = "IBD (Lloyd-Price et al., 2019)",
    CDI = "CDI (Ferretti et al., 2023)",
    CRC = "CRC (Yachida et al., 2019)",
    CRC_ZELLER = "CRC (Zeller et al., 2014)",
    CRC_WIRBEL = "CRC (Wirbel et al., 2018)",
    CRC_VOGTMANN = "CRC (Vogtmann et al., 2016)",
    IBD_FRANZOSA = "IBD (Franzosa et al., 2018)"
  ))


```

```{r}
#| label: tbl-summary-main
#| tbl-cap: "Summary of the number of samples and taxa in each main dataset, stratified by disease status."

tbl_clean |>
  filter(
    dataset %in% c("CDI (Ferretti et al., 2023)", "CRC (Yachida et al., 2019)", "IBD (Lloyd-Price et al., 2019)")
    ) |>
  gt(rowname_col = "group", groupname_col = "dataset") |>
  tab_stub_indent(
    rows = everything(),
    indent = 5
  ) |>
  opt_align_table_header(align = "left")

```

```{r}
#| label: tbl-summary-supp
#| tbl-cap: "Summary of the number of samples and taxa in each main dataset, stratified by disease status."

tbl_clean |>
  filter(
    dataset %in% c("CRC (Zeller et al., 2014)", "CRC (Wirbel et al., 2018)", "CRC (Vogtmann et al., 2016)", "IBD (Franzosa et al., 2018)")
    ) |>
  gt(rowname_col = "group", groupname_col = "dataset") |>
  tab_stub_indent(
    rows = everything(),
    indent = 5
  ) |>
  opt_align_table_header(align = "left")

```

```{r}
# Load datasets

if (reset) {
  
  df_res_main_IBD_flashweave <- process_data("IBD", "flashweave", symmetric = TRUE)
  df_res_main_CDI_flashweave <- process_data("CDI", "flashweave", symmetric = TRUE)
  df_res_main_CRC_flashweave <- process_data("CRC", "flashweave", symmetric = TRUE)
  
  df_res_main_IBD_beem <- process_data("IBD", "beem", symmetric = FALSE)
  # df_res_main_CDI_beem <- process_data("CDI", "beem", symmetric = FALSE)
  df_res_main_CRC_beem <- process_data("CRC", "beem", symmetric = FALSE)
  
  df_res_supp_IBD_franzosa_flashweave <- process_data("IBD_franzosa", "flashweave", symmetric = TRUE)
  df_res_supp_CRC_vogtmann_flashweave <- process_data("CRC_vogtmann", "flashweave", symmetric = TRUE)
  df_res_supp_CRC_wirbel_flashweave <- process_data("CRC_wirbel", "flashweave", symmetric = TRUE)
  df_res_supp_CRC_zeller_flashweave <- process_data("CRC_zeller", "flashweave", symmetric = TRUE)
  
  # Combine all datasets
  df_res_main_flashweave <- bind_rows(df_res_main_IBD_flashweave, df_res_main_CDI_flashweave, df_res_main_CRC_flashweave)
  df_res_main_beem <- bind_rows(df_res_main_IBD_beem, df_res_main_CRC_beem)
  df_res_supp_flashweave <- bind_rows(df_res_supp_IBD_franzosa_flashweave, df_res_supp_CRC_vogtmann_flashweave,
                                      df_res_supp_CRC_wirbel_flashweave, df_res_supp_CRC_zeller_flashweave)

  # Save the final combined datasets
  saveRDS(df_res_main_flashweave, file = str_c(result_dir(), "/df_res_main_flashweave_", today() |> str_remove_all("-"),".rds"))
  saveRDS(df_res_main_beem, file = str_c(result_dir(), "/df_res_main_beem_", today() |> str_remove_all("-"),".rds"))
  saveRDS(df_res_supp_flashweave, file = str_c(result_dir(), "/df_res_supp_flashweave_", today() |> str_remove_all("-"),".rds"))
  
  rm(
    df_res_main_IBD_flashweave, df_res_main_CDI_flashweave, df_res_main_CRC_flashweave,
    df_res_main_IBD_beem, df_res_main_CRC_beem,
    df_res_supp_IBD_franzosa_flashweave, df_res_supp_CRC_vogtmann_flashweave, df_res_supp_CRC_wirbel_flashweave,
    df_res_supp_CRC_zeller_flashweave
  )
  
} else {
  
  df_res_main_flashweave <- 
      readRDS(
      file = 
        list.files(result_dir(), pattern = "df_res_main_flashweave_.*\\.rds$", full.names = TRUE) |> 
        sort() |> 
        tail(1)
      )
  
  df_res_main_beem <-
    readRDS(
      file = 
        list.files(result_dir(), pattern = "df_res_main_beem_.*\\.rds$", full.names = TRUE) |> 
        sort() |> 
        tail(1)
      )
  
  df_res_supp_flashweave <-
    readRDS(
      file =
        list.files(result_dir(), pattern = "df_res_supp_flashweave_.*\\.rds$", full.names = TRUE) |> 
        sort() |> 
        tail(1)
      )
}

```

When examining the CRC dataset from @Yachida2019, we found that net interaction values were significantly higher in healthy samples compared to diseased samples, contrary to the pattern reported by @CorralLopez2026. This result directly opposes the proposed universal shift toward positive interactions in disease. Importantly, this trend was consistent across network inference methods (FlashWeave and BEEM-Static), subsampling fractions, and was further supported by Positive Edge Percentage (PEP) analyses (@fig-main-CRC; Supp. Fig. -@fig-CRC-pep).

```{r}
#| label: fig-main-CRC
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Net interactions results for the CRC dataset from @Yachida2019 across 60, 70 and 80% subsampling on 100 bootstrap iterations. (A) Net interactions results for the CRC dataset using the FlashWeave tool assuming homogeneous samples (`heterogeneous = FALSE`). (B) Net interactions results for the CRC dataset using the BEEM-Static tool with 30% prevalence filtering."
  
p1 <- df_res_main_flashweave |>
  filter(disease == "CRC") |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_boxplot(alpha = 0.4) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  ggh4x::facet_nested(cols = vars(fraction)) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold",
    size = 4
  ) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- df_res_main_beem |>
  filter(disease == "CRC") |>
  ggplot() +
  aes(x = group, y = rho, color = group, fill = group) +
  geom_boxplot(alpha = 0.4) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
  scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
  ggh4x::facet_nested(cols = vars(fraction)) +
  labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
  stat_compare_means(
    comparisons = list(c("Healthy", "Diseased")),
    method = "wilcox.test",
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.8,
    fontface = "bold",
    size = 4
  ) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

(p1 + p2) + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A") & theme(legend.position = "bottom")

```

To assess whether this discrepancy was dataset-specific, we analyzed additional CRC cohorts [@Vogtmann2016; @Wirbel2019 @Zeller2014]. In these datasets, net interactions were generally higher in diseased samples, in agreement with previous reports (@fig-supp-CRC). While group-level differences in net interaction values are observed, the distributions are not clearly separated, particularly in Figure -@fig-supp-CRC A. This suggests that ENBI mainly reflects cohort-level structure rather than independent subject-level variation, since network inference requires multiple samples.

```{r}
#| label: fig-supp-CRC
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Net interactions results for the supplementary CRC dataset across 60, 70 and 80% subsampling on 100 bootstrap iterations using the FlashWeave tool, assuming homogeneous samples (`heterogeneous = FALSE`). (A) Net interactions results for the CRC dataset from @Vogtmann2016. (B) Net interactions results for the CRC dataset from @Wirbel2019. (c) Net interactions results for the CRC dataset from @Zeller2014."

df_res_supp_flashweave |>
    filter(disease == "CRC_vogtmann") |>
    ggplot() +
    aes(x = group, y = rho, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  
df_res_supp_flashweave |>
    filter(disease == "CRC_zeller") |>
    ggplot() +
    aes(x = group, y = rho, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  
df_res_supp_flashweave |>
    filter(disease == "CRC_wirbel") |>
    ggplot() +
    aes(x = group, y = rho, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  
patchwork::plot_layout(ncol = 3, guides = "collect") + plot_annotation(tag_levels = "A") & theme(legend.position = "bottom")


```

In contrast to CRC, both IBD [@LloydPrice2019] and CDI [@Ferretti2023] datasets showed significantly higher net interaction values in diseased samples compared to healthy controls (@fig-main-CDI-IBD). These findings are consistent with the interaction imbalance pattern described by @CorralLopez2026. Similar trends were observed in the independent IBD cohort from @Franzosa2018, and were robust across inference methods (Supp. Figs. -@fig-IBD-supp and -@fig-IBD-beem).

```{r}
#| label: fig-main-CDI-IBD
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Net interactions results for the CDI [@Ferretti2023] and IBD dataset [@LloydPrice2019] across 60, 70 and 80% subsampling on 100 bootstrap iterations. (A) Net interactions results for the IBD dataset using the FlashWeave tool assuming homogeneous samples (`heterogeneous = FALSE`). (B) Net interactions results for the CDI dataset using the FlashWeave tool assuming homogeneous samples (`heterogeneous = FALSE`)."

df_res_main_flashweave |>
    filter(disease == "IBD") |>
    ggplot() +
    aes(x = group, y = rho, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  
df_res_main_flashweave |>
    filter(disease == "CDI") |>
    ggplot() +
    aes(x = group, y = rho, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +

patchwork::plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A") & theme(legend.position = "bottom")

```


## Conclusion

In this work, we reassessed the relationship between microbial interaction imbalance and disease status across multiple gut microbiome datasets. While increased net positive interactions in diseased samples were observed in IBD and CDI, this pattern was not consistent across all conditions. Notably, the CRC dataset from @Yachida2019 exhibited the opposite trend, with higher net interaction values in healthy samples, a result that was robust across inference methods and subsampling strategies. Consequently, these findings do not support the claim that microbial interaction imbalance constitutes a universal signature of disease, but rather suggest that such metrics are context-dependent.

## References

::: {#refs}
:::

## Supplementary Figures {.unnumbered}

```{r}
#| label: fig-CRC-pep
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "PEP results for the CRC dataset from @Yachida2019 on 100 bootstrap iterations. (A) PEP results for the CRC dataset using the FlashWeave tool assuming homogeneous samples (`heterogeneous = FALSE`) across 60, 70 and 80% subsampling. (B) PEP results for the CRC dataset using the BEEM-Static tool with a 30% prevalence filtering across 70, 80 and 90% subsampling ."

# CRC with PEP

df_res_main_flashweave |>
    filter(disease == "CRC") |>
    ggplot() +
    aes(x = group, y = pep, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) + 
  
  df_res_main_beem |>
    filter(disease == "CRC") |>
    ggplot() +
    aes(x = group, y = pep, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) + 
  
patchwork::plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A") & theme(legend.position = "bottom")

```

```{r}
#| label: fig-IBD-supp
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Net interactions results for the IBD dataset from @Franzosa2018 across 60, 70 and 80% subsampling on 100 bootstrap iterations using the FlashWeave tool assuming homogeneous samples (`heterogeneous = FALSE`)."


# IBD dataset from @Franzosa2018

df_res_supp_flashweave |>
    filter(disease == "IBD_franzosa") |>
    ggplot() +
    aes(x = group, y = pep, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  )

```

```{r}
#| label: fig-IBD-beem
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Net interactions results for the IBD dataset from @Franzosa2018 across 70, 80 and 90% subsampling on 100 bootstrap iterations using the BEEM-Static tool with a 30% prevalence filtering."


# IBD dataset main with BEEM-Static

df_res_main_beem |>
    filter(disease == "IBD") |>
    ggplot() +
    aes(x = group, y = pep, color = group, fill = group) +
    geom_boxplot(alpha = 0.4) +
    geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    scale_fill_manual(values  = c("Healthy" = "#0078B9", "Diseased" = "#EA0017")) +
    ggh4x::facet_nested(cols = vars(fraction)) +
    labs(x = "", y = "Net interactions", color = "Group", fill = "Group") +
    stat_compare_means(
      comparisons = list(c("Healthy", "Diseased")),
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.8,
      fontface = "bold",
      size = 4
    ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  )


```

