---
title: "Data preprocessing for dysbiotic cohorts"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: true
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

In this section, we will first import the relevant datasets from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI), and then perform several descriptive analyses for each of those datasets. The analyses will include: data cleaning, taxonomic aggregation (if necessary) and rarefaction. 

We will transform each dataset into a `phyloseq` object, before merging them into a `MultiAssayExperiment` object in the last section. For this we need:

- abundance data, a matrix of relative abundance with samples as rows and features (taxa) as columns.
- taxonomy data, a data frame with features (taxa) as rows and taxonomic levels as columns.
- samples data, a data frame with samples as rows and metadata variables as columns.

> TODO: Create dictionaries for metadata variables for each dataset, to keep track of the meaning of each variable and the values they can take, previous and new name, and if it is kept for further analysis or not.

## IBD dataset (Lloyd-Price et al. 2019)

The inflammatory bowel disease (IBD) dataset was obtained from the [IBDMDB](https://ibdmdb.org/), with the results published in Lloyd-Price et al. (2019). The authors of the Science paper did not provide information to which study name the IBD dataset comes from, so the HMP2 study was chosen, which seems like a reasonable choice since it is the most recent study on this page. 

Stool samples were initially collected from 132 participants at multiple time points. Specifically, samples were collected at baseline (that is after the initial colonoscopy, which was used to stratify participants into healthy controls, Crohn's disease (CD) patients, and ulcerative colitis (UC) patients), and then every 2 weeks for 1 years.

Metagenomics (MGX) were performed on 1638 stool samples; DNA was quantified, normalized, and used to prepare sequencing libraries with the Nextera XT kit. Libraries were pooled, quality-checked, and sequenced on an Illumina HiSeq platform to generate ~10 million paired-end reads per sample. Raw sequencing data was then demultiplexed and converted to standard file formats (BAM and FASTQ files) using the Picard suite.

For the analyses, we have the following datasets at our disposal:

- `hmp2_metadata.csv`: contains the metadata for the samples.

From Metagenomics (MGX) and products sections, we have the following file available:

- `taxonomic_profiles.biom.gz`: contains the taxonomic profiles and **relative abundance metagenomics data** for **stool samples**, in BIOM format.

### Importing raw data

In the chunks below, we import the metadata and taxonomic profile for the IBD dataset.

```{r}
# Importing the metadata

metadata <- read_csv(file.path(data_dir(), "01-raw-data/01-IBD", "hmp2_metadata_2018-08-20.csv"), col_types = cols(.default = "c"))

# Importing the taxonomic profiles

data <- read_biom(file.path(data_dir(), "01-raw-data/01-IBD", "taxonomic_profiles.biom"))

```

We document the number of samples and feature in the original dataset.

```{r}

n_samples <- nrow(metadata)
n_features <- ncol(data)

```

There are `r n_samples` samples and `r n_features` features in the original dataset.

### Cleaning taxonomic annotations

For abundance data, we need to filter the taxonomic profiles to keep only the bacterial species relevant for our analyses, and to remove the taxonomic levels from the taxonomic names by keeping only the last available level.

```{r}

df <- data |> biom_data() |> as.matrix() |> t() |> as.data.frame() 
abundance <- df |> select(contains("k__Bacteria"))

names(abundance) <- str_extract(names(abundance), "[^|]+$")

# df |> pivot_longer(everything(), names_to = "feature", values_to = "rel_ab") |> 
#   filter(rel_ab == 1, feature != "k__Bacteria") |>
#   gt(caption = "Features with relative abundance of 1")

```

For taxonomy data, we need to extract the taxonomic levels and names from the taxonomic profiles, and to keep only the last available level for each feature, saved as `taxon_names`, with a letter indicating the taxonomic level between parentheses (e.g., "g" for genus, "f" for family, etc.). Additionally, we also filter the dataset to only keep the "Bacteria" kingdom, since we are only interested in bacterial taxa for our analyses.

```{r}

taxonomy <- 
  tibble(full_name = names(df)) |>
  mutate(
    kingdom = str_extract(full_name, "k__[^|]+"),
    phylum = str_extract(full_name, "p__[^|]+"),
    class = str_extract(full_name, "c__[^|]+"),
    order = str_extract(full_name, "o__[^|]+"),
    family = str_extract(full_name, "f__[^|]+"),
    genus = str_extract(full_name, "g__[^|]+"),
    species = str_extract(full_name, "s__[^|]+")
  ) |>
  mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, kingdom
    )
  ) |>
  mutate(
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(kingdom, "k__"), " (k)")
    )
  ) |>
  filter(kingdom == "k__Bacteria") |>
  select(-full_name)

```

::: {.callout-note}
In the supplementary code provided, the authors filtered out taxa with `_unclassified` annotation. Also, they are also `_nonname` taxa, but they did not filter those out. I filter out both `_unclassified` and `_noname` taxa, but I am not sure if this is the right choice. *Should we filter out both `_unclassified` and `_noname` taxa, or only `_unclassified` taxa?*
:::

We document below how many features have `_unclassified` or `_noname` annotation at each taxonomic level.

```{r}

taxonomy |>
  summarise(across(c(kingdom, phylum, class, order, family, genus, species), list(
    unclassified = ~ sum(str_detect(., "_unclassified"), na.rm = TRUE),
    noname = ~ sum(str_detect(., "_noname"), na.rm = TRUE)
  ))) |>
  pivot_longer(everything(),
               names_to = c("level", "type"),
               names_sep = "_(?=unclassified|noname)",
               values_to = "n") |>
  mutate(level = factor(level, levels = c("kingdom", "phylum", "class", "order", "family", "genus", "species"))) |>
  ggplot() +
  aes(x = type, y = fct_rev(level), fill = n) +
  geom_tile() +
  geom_label(aes(label = n), size = 4) +
  scale_fill_gradient(low = "white", high = "dodgerblue") +
  labs(x = NULL, y = NULL, fill = "Count",
       subtitle = "Unclassified / noname annotations per taxonomic level")

```

For taxa samples, several taxa have the same taxonomic annotation, so we need to aggregate the abundance data by taxonomic annotation, and to keep only one row for each taxon in the taxonomy data.

The duplicates are documented below, with the taxonomic annotation and the number of features with this annotation.

```{r}

taxonomy |> group_by(taxon_name) |> summarise(n = n()) |> filter(n > 1) |>
  gt(caption = "Taxonomic annotation with duplicates") |>
  tab_options(
    container.height = px(400)
  )

```

Since abundance data is in relative scale, we can simply sum the relative abundance of features with the same taxonomic annotation to get the relative abundance of the taxon.

```{r}

abundance <-
  abundance |>
  t() |>
  as.data.frame() |>
  bind_cols(taxonomy |> select(taxon)) |>
  group_by(taxon) |>
  summarise(across(everything(), sum)) |>
  column_to_rownames("taxon") |>
  t() |>
  as.data.frame()

```

We also need to filter the taxonomy data to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <- 
  taxonomy |> 
  distinct(taxon, .keep_all = TRUE) |> 
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

Since the IBD dataset contains higher level taxonomic annotations in addition to species level annotations, we can also create a `phyloseq` object with only species level annotations, by filtering the taxonomy data to keep only species level annotations, and modify the abundance data accordingly.

Additionally, we also filter out species for which `_unclassified` or `_noname` annotation is present.

```{r}

taxonomy <- 
  taxonomy |> 
  rownames_to_column("taxon") |>
  filter(!is.na(species)) |> 
  column_to_rownames("taxon") |>
  filter(!str_detect(species, "_unclassified|_noname"))

```

### Documenting abundance distribution

```{r}

abundance <- 
  abundance |>
  t() |>
  as.data.frame() |>
  rownames_to_column("taxon") |>
  filter(taxon %in% rownames(taxonomy)) |> 
  column_to_rownames("taxon") |>
  mutate(across(everything(), as.numeric)) |>
  t() |>
  as.data.frame()

```

We check if samples rows sum to 1.

```{r}

row_sums <- rowSums(abundance)

if (all(row_sums == 1)) {
  cat("All samples rows sum to 1.\n")
} else {
  cat("Not all samples rows sum to 1.\n")
}

row_sums |> summary()

```

We renormalize the abundance data to make sure that all samples rows sum to 1 and discard samples for which the sum of relative abundance is 0 (i.e., samples for which we do not have any bacterial species annotation after filtering out unclassified and noname species).

```{r}

dropped <- rownames(abundance)[rowSums(abundance) == 0]
cat("Dropped samples:\n"); print(dropped)

abundance <- abundance[rowSums(abundance) != 0, ] |> (\(x) x / rowSums(x))()

rowSums(abundance) |> summary()

```

We check if there are any singletons (i.e., taxa that are present in only one sample) in the dataset, and if so, we document them by showing the taxonomic annotation and the number of samples with this annotation.

```{r}

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  filter(rel_ab > 0.05) |>
  group_by(taxon) |>
  mutate(singleton = ifelse(n() == 1, "Singleton", "Non-singleton")) |>
  ungroup() |>
  filter(singleton == "Singleton") |>
  ggplot() +
  aes(x = taxon, y = sample_id, fill = rel_ab) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue", labels = scales::label_percent(accuracy = 1)) +
  labs(x = "Taxon", y = "Sample ID", fill = "Relative abundance",
       subtitle = "Singleton taxa where rel_ab > 5% (present in only one sample)") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

::: {.callout-note}
The singletons in the dataset should not pose a problem for the downstream analyses, since they have a low relative abundance.
:::

We also document the distribution of the number of taxa per sample, as well as the mean relative abundance across samples over all taxa.

```{r}

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  filter(rel_ab > 0) |>
  group_by(sample_id) |>
  summarise(n_taxa = n()) |>
  ggplot() +
  aes(x = n_taxa) +
  geom_histogram(bins = 30) +
  labs(x = "Number of taxa", y = "Number of samples",
       subtitle = "Distribution of the number of taxa per sample") +

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  filter(rel_ab > 0) |>
  group_by(sample_id) |>
  summarise(mean_rel_ab = mean(rel_ab)) |>
  ggplot() +
  aes(x = mean_rel_ab) +
  geom_histogram(bins = 30) +
  labs(x = "Mean relative abundance", y = "Number of samples",
       subtitle = "Distribution of the mean relative abundance across samples") +

  plot_layout(nrow = 2)

```
Below, we check if there are any samples with a taxon with relative abundance of 1, and if so, we document them by showing the sample ID and the taxonomic annotation of the taxon with relative abundance of 1.

```{r}

abundance |>
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_ab") |>
  filter(rel_ab == 1) |>
  gt(caption = "Samples with a taxon with relative abundance of 1")

```


::: {.callout-warning}
A sample has a taxon with relative abundance of 1. This means that this sample has only one taxon with non-zero relative abundance. However, this taxon was not identified above (before the renormalization), so this is likely due the renormalization step.
:::

### Documenting available metadata

For samples data, we need to filter the metadata to keep only the samples for which we have metagenomics data, and to set the sample IDs as row names. Additionally, since not all covariates in the metadata are relevant for our analyses, we also select only a subset of the metadata variables to keep in the samples data. 

We document (some) of the original metadata variables in the plots below.

```{r}
# we filter out columns with only NA values

metadata_filtered <- metadata |> filter(data_type == "metagenomics") |> select(where(~!all(is.na(.x))))
samples <- metadata |> clean_names() |> filter(data_type == "metagenomics") |> select(where(~!all(is.na(.x))))

```

::: panel-tabset

#### Age at diagnosis
```{r}
#| results: asis
samples |>
  filter(!is.na(age_at_diagnosis)) |>
  ggplot() +
  aes(x = diagnosis, y = age_at_diagnosis |> as.numeric(), fill = diagnosis, color = diagnosis) +
  geom_violin(alpha = 0.4, color = NA) + 
  geom_point(position = position_jitter(width = 0.2), size = 1, alpha = 0.6) +
  scale_fill_manual(values = c("nonIBD" = "#0078B9", "CD" = "#EA0017", "UC" = "#FF7F0E")) +
  scale_color_manual(values = c("nonIBD" = "#0078B9", "CD" = "#EA0017", "UC" = "#FF7F0E")) +
  labs(x = "", y = "Age", fill = "Diagnosis", color = "Diagnosis",
       subtitle = "Distribution of age across diagnosis groups")
```

#### Education level
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = education_level, fill = diagnosis) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("nonIBD" = "#0078B9", "CD" = "#EA0017", "UC" = "#FF7F0E")) +
  labs(y = "", x = "Education level", fill = "Diagnosis",
       subtitle = "Distribution of education level \nacross diagnosis groups") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


#### Occupation
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = occupation, fill = diagnosis) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("nonIBD" = "#0078B9", "CD" = "#EA0017", "UC" = "#FF7F0E")) +
  labs(y = "", x = "Occupation", fill = "Diagnosis",
       subtitle = "Distribution of occupation across diagnosis groups") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#### Diagnosis
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = diagnosis, fill = diagnosis) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("nonIBD" = "#0078B9", "CD" = "#EA0017", "UC" = "#FF7F0E")) +
  labs(y = "", x = "", fill = "Diagnosis",
       subtitle = "Distribution of diagnosis groups")

```
:::

::: {.callout-note}
There is no age data for the healthy group, since healthy participants were not diagnosed with IBD, so they do not have an age at diagnosis.
:::

The selected variables are: "Project", "External ID", "Participant ID", "data_type", "week_num", "visit_num", "IntervalName", and "diagnosis".

```{r}

samples <- 
  metadata |> 
  select("External ID", "Participant ID", "data_type", "week_num", "visit_num", "IntervalName", "diagnosis") |>
  clean_names() |>
  filter(data_type == "metagenomics") |>
  column_to_rownames("external_id")

samples |> glimpse()

```

We also filter out the dropped samples (for which the sum of relative abundance is 0) from the samples data.

```{r}

samples <- 
  samples |> 
  rownames_to_column("sample_id") |>
  filter(!sample_id %in% dropped) |>
  column_to_rownames("sample_id")

```


```{r}
#| fig-height: 4
#| fig-width: 15

samples |>
  rownames_to_column("sample_id") |>
  select(sample_id, participant_id, visit_num, week_num, interval_name, diagnosis) |>
  group_by(visit_num, participant_id) |>
  mutate(total_samples = n()) |>
  ggplot() +
  aes(x = participant_id, fill = diagnosis) +
  geom_bar() +
  labs(x = "Participant ID", y = "Number of samples", fill = "Diagnosis",
       subtitle = "Number of samples per participant and diagnosis") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

#### Creating a dictionary for metadata variables

We create a dictionary for the metadata variables, to keep track of the meaning of each variable and the values they can take, previous and new name, and if it is kept for further analysis or not.

```{r}

IBD_dictionary <-
  tibble(
    original_variable = metadata_filtered |> names(),
    new_variable = metadata_filtered |> clean_names() |> names(),
    description = c(
      "Project name",
      "External ID of the sample",
      "Participant ID",
      "Site and sub-collection identifier",
      "Type of data (e.g., metagenomics, metabolomics)",
      "Week number of sample collection",
      "Date sample was received",
      "Number of days since previous sample",
      "Visit number of sample collection",
      "Associated research project name",
      "PDO number (project/delivery order)",
      "GSSR IDs associated with the sample",
      "Sequencing product used",
      "Library construction set identifier",
      "Aggregated sequencing lanes",
      "Work request ID",
      "Number of lanes included in aggregation",
      "Total raw reads",
      "Reads passing quality filters",
      "Reads failing QC",
      "Reads mapping to human genome",
      "Interval name (e.g., baseline, week 2)",
      "Sequence order of the interval",
      "Project-specific sample ID",
      "Name of collection site",
      "Age at IBD diagnosis (years)",
      "Highest education level attained",
      "Participant occupation",
      "Whole blood received at Broad Institute",
      "Serum tube 1 received at CSMC",
      "Serum tubes 2–4 received at MGH",
      "Flora sample received at MGH",
      "RNA/DNA sample received at Broad Institute",
      "ECP sample received at Washington University",
      "Age at consent (years)",
      "Diagnosis group (e.g., nonIBD, CD, UC)",
      "Frequency of sugary soft drinks, tea, or coffee",
      "Frequency of diet drinks with artificial sweeteners",
      "Frequency of fruit juice consumption",
      "Frequency of water consumption",
      "Frequency of alcohol consumption",
      "Frequency of probiotic foods (yogurt, kefir, sauerkraut)",
      "Frequency of dairy consumption",
      "Probiotic supplement use",
      "Frequency of fruit consumption (no juice)",
      "Frequency of vegetable consumption",
      "Medications received in the past 2 weeks",
      "Frequency of bean/legume consumption",
      "Frequency of whole grain consumption",
      "Frequency of refined starch consumption",
      "Frequency of egg consumption",
      "Frequency of processed meat consumption",
      "Frequency of red meat consumption",
      "Frequency of white meat consumption",
      "Frequency of shellfish consumption",
      "Frequency of fish consumption",
      "Frequency of sweets consumption",
      "Antibiotic use in past 2 weeks",
      "Chemotherapy in past 2 weeks",
      "Immunosuppressant use in past 2 weeks",
      "Colonoscopy or procedure in past 2 weeks",
      "Oral contrast use in past 2 weeks",
      "Diarrhea in past 2 weeks",
      "Hospitalization in past 2 weeks",
      "History of bowel surgery",
      "Frequency of unsweetened tea or coffee",
      "Fecal calprotectin level (ng/mL)",
      "Presence of uveitis (extraintestinal manifestation)",
      "Presence of erythema nodosum",
      "Presence of aphthous ulcers",
      "Presence of pyoderma gangrenosum",
      "Presence of anal fissure",
      "Presence of new fistula",
      "Presence of abscess",
      "Whether HBI was completed",
      "HBI: general wellbeing score",
      "HBI: abdominal pain score",
      "HBI: number of liquid or soft stools in past 24 hours",
      "HBI: abdominal mass score",
      "HBI: arthralgia score",
      "Harvey-Bradshaw Index (HBI) total score",
      "Hispanic or Latino origin",
      "History of breast cancer",
      "History of cholangiocarcinoma",
      "History of colon or rectal cancer",
      "History of Hodgkin's lymphoma",
      "History of liver cancer",
      "History of lung cancer",
      "History of lymphoma (NOS)",
      "History of non-Hodgkin's lymphoma",
      "History of ovarian cancer",
      "History of prostate cancer",
      "History of celiac disease",
      "History of chronic bronchitis",
      "History of dermatitis herpetiformis",
      "History of familial Mediterranean fever",
      "History of Graves' disease",
      "History of Guillain-Barré syndrome",
      "History of Hashimoto's thyroiditis",
      "History of idiopathic pulmonary fibrosis",
      "History of idiopathic thrombocytopenic purpura",
      "History of irritable bowel syndrome",
      "History of alopecia areata",
      "History of multiple sclerosis",
      "History of myasthenia gravis",
      "History of myocarditis",
      "History of neuropathy",
      "History of pericarditis",
      "History of pemphigus vulgaris",
      "History of pernicious anemia",
      "History of polymyositis or dermatomyositis",
      "History of primary biliary cirrhosis",
      "History of primary sclerosing cholangitis",
      "History of ankylosing spondylitis",
      "History of psoriasis",
      "History of rheumatoid arthritis",
      "History of sarcoidosis",
      "History of scleroderma / systemic sclerosis",
      "History of Sjögren's syndrome",
      "History of systemic lupus erythematosus",
      "History of temporal arteritis",
      "History of thyroid disease (uncertain diagnosis)",
      "History of type I (juvenile) diabetes",
      "History of vitiligo",
      "History of arthritis (uncertain diagnosis)",
      "History of Wegener's granulomatosis",
      "History of asthma",
      "History of autoimmune hemolytic anemia",
      "History of autoimmune hepatitis",
      "History of Behçet's syndrome",
      "Whether patient signed informed consent",
      "Whether subject withdrew from the study",
      "Whether subject was terminated by investigator",
      "Reason for withdrawal or termination",
      "Partial endoscopy performed",
      "Baseline Montreal classification location",
      "Explanation for 'other' responses",
      "Self-reported race",
      "Presence of arthralgias (joint pain)",
      "Presence of pyoderma gangrenosum (follow-up)",
      "Presence of erythema nodosum (follow-up)",
      "Presence of uveitis (follow-up)",
      "Whether SCCAI was completed",
      "SCCAI: bowel frequency during the day",
      "SCCAI: bowel frequency during the night",
      "SCCAI: urgency of defecation",
      "SCCAI: blood in the stool",
      "SCCAI: general well-being in past 24 hours",
      "Simple Clinical Colitis Activity Index (SCCAI) total score",
      "Partial endoscopy performed (follow-up)",
      "Biological sex",
      "Specified race (free text)",
      "Whether stool sample was collected",
      "Tube A collected for DNA/RNA analysis",
      "Tube A collected for metabolomics",
      "Tube A collected for storage",
      "Tube A collected for viromics",
      "Tube B collected for fecal calprotectin",
      "Fecal calprotectin measurement",
      "Tube B collected for proteomics",
      "Viromics sample received at Baylor",
      "Fecal calprotectin sample received at MGH",
      "Proteomics sample received at LBNL",
      "Stool sample ID for tube A (EtOH)",
      "Sample ID for tube B (no preservative)",
      "Tubes A and B received at Broad Institute"
    ),
    removed = c(
      TRUE, FALSE, FALSE, TRUE, TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE,
      TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE
    )
  ) |>
  mutate(new_variable = case_when(
    new_variable == "external_id" ~ "uid",
    new_variable == "participant_id" ~ "pid",
    TRUE ~ new_variable
  ))

IBD_dictionary |> gt(caption = "Dictionary for the metadata variables in the IBD dataset") |>
  tab_options(container.height = px(800))

```


```{r}

samples <-
  samples |> 
  rownames_to_column("sample_id") |>
  dplyr::rename(uid = sample_id, pid = participant_id) |>
  column_to_rownames("uid")

samples |> glimpse()

```

### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibd <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibd

```

## IBS dataset (Mars et al., 2020)

The irritable bowel syndrome (IBS) dataset in Mars et al. (2020) was obtained from [Mendeley data](https://data.mendeley.com/datasets/29n2z5r5ph/3) plateform. 

Metagenomics were performed for 474 stool samples recolted from 77 participants, separated into healthy controls (H),  constipation predominant (IBS-C), diarrhea predominant (IBS-D), or mixed (IBS-M). DNA was extracted from stool samples using the QIAGEN PowerSoil kit and quantified by NanoDrop and PicoGreen assays. Shotgun metagenomic libraries were prepared using a modified Nextera XT protocol and sequenced on either a HiSeq 2500 or NextSeq platform with single-end reads. Samples from three cohorts were randomly distributed across two sequencing runs to reduce batch effects. Reads were trimmed to 100 bp and aligned to a prokaryotic reference genome collection at 97% identity using BURST, with low-depth samples (< 10,000 reads) removed. Functional profiles were generated using KEGG Orthology annotations, and low-abundance gene estimates were further improved using SHOGUN.

For the analyses, we have the following datasets at our disposal:

- `metagenomics.xlsx`: contains the **metagenomics count data** and metadata for biopsy samples, in CSV format. This correspond to the file "Supplementary data I Microbiome data tables stool and biopsy.xlsx" in the plateform (version 3 from 2020-03-11).

### Importing raw data

In the chunk below, we import the metadata and abundance data for the IBS dataset.

```{r}

data_biopsy <- readxl::read_excel(file.path(data_dir(), "01-raw-data/02-IBS", "metagenomics.xlsx"), sheet = 2)
metadata <- readxl::read_excel(file.path(data_dir(), "01-raw-data/02-IBS", "metagenomics.xlsx"), sheet = 5)

```

We first document the number of samples and features in the original dataset.

```{r}

n_samples <- nrow(metadata)
n_features <- ncol(data_biopsy) - 1 # we remove the first column

```

There are `r n_samples` samples and `r n_features` features in the original dataset.

### Cleaning taxonomic annotations

We then transform the abundance data into a `phyloseq` object in the next steps, but before that, we need to perform some data cleaning. For this, we need to filter the abundance data to keep only bacterial species, and to remove the taxonomic levels from the taxonomic names by keeping only the last available level.

```{r}

abundance <- data_biopsy |> filter(str_detect(`#OTU ID`, "k__Bacteria"))

```

For taxonomy data, we need to extract the taxonomic levels and names from the taxonomic profiles (OTU column in `data_biopsy`), and to keep only the last available level for each feature, saved as `taxon_names`, with a letter indicating the taxonomic level between parentheses (e.g., "g" for genus, "f" for family, etc.).

```{r}

taxonomy <-
  tibble(full_name = abundance$`#OTU ID`) |>
  mutate(
    kingdom = str_extract(full_name, "k__[^;]+"),
    phylum = str_extract(full_name, "p__[^;]+"),
    class = str_extract(full_name, "c__[^;]+"),
    order = str_extract(full_name, "o__[^;]+"),
    family = str_extract(full_name, "f__[^;]+"),
    genus = str_extract(full_name, "g__[^;]+"),
    species = str_extract(full_name, "s__[^;]+")
  ) |>
    mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, kingdom
    )
  ) |>
  mutate(
    across(c(species, genus, family, order, class, phylum, kingdom), 
           ~ str_remove_all(., "\\[|\\]")),
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(kingdom, "k__"), " (k)")
    )
  ) |>
  select(-full_name)

```

We check if there are any missing species annotations.

```{r}

cat("There are", nrow(taxonomy |> filter(is.na(species)) == "TRUE"), "features without species annotation.\n")

```

We additionally check if there are any duplicated taxonomic annotations, and if so, we document them by showing the taxonomic annotation and the number of features with this annotation. The duplicates will be merged in taxonomy, as well as in abundance data.

```{r}

taxonomy |> 
  group_by(taxon_name) |> 
  summarise(n = n()) |> 
  filter(n > 1) |>
  gt(caption = "Taxonomic annotation with duplicates")

```

To merge duplicates in abundance, we can simply sum the counts of features with the same taxonomic annotation to get the count of the taxon.

```{r}

abundance <-
  abundance |>
  bind_cols(taxonomy |> select(taxon)) |>
  group_by(taxon) |>
  summarise(across(-`#OTU ID`, sum)) |>
  column_to_rownames("taxon")

abundance |> head() |> select(1:5)

```


### Documenting abundance distribution

We check if there are any singletons (i.e., taxa that are present in only one sample) in the dataset, and if so, we document them by showing the taxonomic annotation and the number of samples with this annotation.

```{r}
#| fig-height: 4
#| fig-width: 6

abundance |> 
  rownames_to_column("taxon") |>
  pivot_longer(-taxon, names_to = "sample_id", values_to = "counts") |>
  group_by(sample_id) |>
  mutate(rel_ab = counts / sum(counts)) |>
  ungroup() |>
  select(-counts) |>
  filter(rel_ab > 0.05) |>
  group_by(taxon) |>
  mutate(singleton = ifelse(n() == 1, "Singleton", "Non-singleton")) |>
  ungroup() |>
  filter(singleton == "Singleton") |>
  ggplot() +
  aes(x = taxon, y = sample_id, fill = rel_ab) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue", labels = scales::label_percent(accuracy = 1)) +
  labs(x = "Taxon", y = "Sample ID", fill = "Relative abundance",
       subtitle = "Singleton taxa where rel_ab > 5% (present in only one sample)") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

We also document the distribution of the number of taxa per sample, as well as the mean counts across samples over all taxa.

```{r}

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  mutate(rel_ab = count / sum(count)) |>
  ungroup() |>
  filter(rel_ab > 0) |>
  group_by(sample_id) |>
  summarise(n_taxa = n()) |>
  ggplot() +
  aes(x = n_taxa) +
  geom_histogram(bins = 30) +
  labs(x = "Number of taxa", y = "Number of samples",
       subtitle = "Distribution of the number of taxa per sample") +

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  summarise(mean_count = mean(count)) |>
  ggplot() +
  aes(x = mean_count |> log10()) +
  geom_histogram(bins = 30) +
  labs(x = "Mean counts", y = "Number of samples",
       subtitle = "Distribution of the mean counts across samples") +

  plot_layout(nrow = 2)

```

We filter duplicates in taxonomy to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <- 
  taxonomy |> 
  distinct(taxon, .keep_all = TRUE) |> 
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

### Documenting available metadata

For sample (metadata) data, we will only keep relevant metadata variables for our analyses, and we set the sample IDs as row names. We document the original metadata variables below.

```{r}

samples <- metadata |> clean_names()

```

::: panel-tabset

#### Age
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = cohort, y = age, color = cohort, fill = cohort) +
  geom_violin(alpha = 0.6, color = NA) +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.8) +
  scale_color_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  scale_fill_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  labs(x = "", y = "Age", color = "Cohort", fill = "Cohort",
       subtitle = "Distribution of age across cohorts") +
  theme(
    axis.text.x = element_blank()
  )
```

#### BMI
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = cohort, y = bmi, color = cohort, fill = cohort) +
  geom_violin(alpha = 0.6, color = NA) +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.8) +
  scale_color_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  scale_fill_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  labs(x = "", y = "BMI", color = "Cohort", fill = "Cohort",
       subtitle = "Distribution of BMI across cohorts") +
  theme(
    axis.text.x = element_blank()
  )
```

#### Gender
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = gender, fill = cohort) +
  geom_bar(position = "dodge") +
  scale_fill_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  labs(x = "", y = "Count", color = "Cohort", fill = "Cohort",
       subtitle = "Distribution of gender across cohorts")
```

#### Cohort
```{r}
#| results: asis
samples |>
  ggplot() +
  aes(x = cohort, fill = cohort) +
  geom_bar(position = "dodge") +
  scale_fill_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  labs(x = "", y = "Count", color = "Cohort", fill = "Cohort",
       subtitle = "Distribution of samples across cohorts") +
   theme(
    axis.text.x = element_blank()
  )
```
:::

```{r}

samples <-
  metadata |>
  clean_names() |>
  mutate(group = ifelse(cohort %in% c("C", "D"), "IBS", cohort),
         group = group |> factor()) |>
  mutate(time = str_replace_all(timepoint, c("1st" = "1", "2nd" = "2")),
         time = time |> factor())

```


```{r}

samples |>
  group_by(study_id, cohort) |>
  summarize(n_samples = n()) |>
  ungroup() |>
  ggplot() +
  aes(x = study_id |> factor(), y = n_samples, fill = cohort) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("H" = "#0078B9", "C" = "#EA0017", "D" = "#EA8777"),
    labels = c("H" = "Healthy", "C" = "IBS-C", "D" = "IBS-D")
    ) +
  scale_y_continuous(breaks = scales::breaks_width(1)) +
  labs(x = "Subject ID", y = "Number of samples",
       subtitle = "Number of samples per subject")

```

#### Creating a dictionary for metadata variables

We construct below a dictionary to keep track of the metadata variables, their meaning, the values they can take, and if they are kept for further analyses or not.

```{r}

dictionary_ibs <- 
  tibble(
  orginal_variables = metadata |> names(),
  new_variables = c("uid", "pid", "time", "id_on_tube", "group", "age", "bmi", "gender"),
  description =
  c("Unique identifier for each sample",
    "Participant ID",
    "Time points of the sample collection, originally encoded as '1st' and '2nd', renamed '1' and '2'",
    "ID on tube",
    "Cohort to which the sample belongs, with three levels: 'H' for healthy controls, 'D' for IBS patients with predominantly diarrhea stool form, and 'C' for IBS patients with predominantly constipation stool form. For analyses, 'C' and 'D' levels will be merged into a single level 'IBS'.",
    "Age of the participant at the time of sample collection",
    "Body mass index (BMI) of the participant at the time of sample collection (calculated as weight in kg divided by height in m squared)",
    "Self-declared gender, with two levels male (M) and female (F)"
    ),
  removed = 
  c(FALSE, FALSE, FALSE, TRUE, FALSE, TRUE, TRUE, TRUE)
  )

dictionary_ibs |> gt(caption = "Dictionary of metadata variables for the IBS dataset")

```


```{r}

samples <-
  samples |>
  select(sample_id, study_id, time, cohort, group) |>
  dplyr::rename(
    uid = sample_id,
    pid = study_id
  ) |>
  column_to_rownames("uid")

samples |> glimpse()

```


### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibs <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibs

```

## CRC dataset (Yachida et al., 2019)

The colorectal cancer (CRC) dataset in Yashida et al. (2019) was obtained from the DDBJ Sequence Read Archive (DRA) under the names [DRA006684](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA006684) and [DRA008156](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA008156); however, those are raw sequencing data, and the authors mentioned having used only processed datasets.

Processed datasets seems to be available as [Supplementary Tables](https://www.nature.com/articles/s41591-019-0458-7#Sec23), therefore these will be used for the subsequent analyses.

Stool samples were collected after bowel prep and stored on dry ice. DNA was extracted using bead-beating with a GNOME kit, quality-checked by TapeStation, and stored at −80°C. Sequencing libraries were prepared with the Nextera XT kit and verified for quality before paired-end whole-genome shotgun sequencing on an Illumina HiSeq 2500, generating 150 bp reads with a target output of 5 Gb per sample.

The samples were classified into nine groups according to colonoscopic and histological findings: 

- normal (no remarkable colonoscopic findings) and a few polyps (up to two small (<5 mm) polyps) were classified as "Healthy";
- multiple polypoid adenomas with low-grade dysplasia and polypoid adenoma(s) with high-grade dysplasia were classified as "MP";
- stage 0 CRC was classified as "Stage_0"; 
- stage I CRC and stage II CRC were classified as "Stage_I_II"; 
- stage III CRC and stage IV CRC were classified as "Stage_III_IV". 
- the remaining group was normal with a history of colorectal surgery, and was classified as "HS".

We have the following datasets at our disposal:

- `41591_2019_458_MOESM3_ESM.xlsx`: contains the processed datasets for **stool samples**, in Excel format, which contains participants metadata and **metagenomics relative abundance data**. This correspond to the file "Supplementary data 3.xlsx" in the article (version 1 from 2019-11-28).

### Importing raw data

We import those datasets in the chunks below.

```{r}

data <- read_xlsx(file.path(data_dir(), "01-raw-data/03-CRC", "41591_2019_458_MOESM3_ESM.xlsx"), skip = 3, sheet = 13)

metadata <- read_xlsx(file.path(data_dir(), "01-raw-data/03-CRC", "41591_2019_458_MOESM3_ESM.xlsx"), skip = 2, sheet = 4)

```

We first document the number of samples and features in the original dataset.

```{r}

n_samples <- nrow(metadata)
n_features <- ncol(data) - 1 # we remove the first column

```

There are `r n_samples` samples and `r n_features` features in the original dataset.

As in the previous datasets, we need to perform some data cleaning before creating the `phyloseq` object. For abundance data, we need to filter the dataset to keep only bacterial species.

First, we notice that the number of samples in metadata and abundance data do not match, so we need to filter the abundance data to keep only samples for which we have metadata.

```{r}

abundance <- data |> select(any_of(metadata$Subject_ID))

```

Now, we see that we have `r nrow(metadata)`samples in metadata, and `r ncol(abundance)` samples in abundance data, so there is still a mismatch between the two datasets. We document this below.

```{r}

setdiff(metadata$Subject_ID, colnames(abundance))

```

::: {.callout-note}
After checking manually, we see that this sample is present in the metadata, but not in the abundance data, so we can simply filter it out from the metadata.
:::

```{r}

metadata <- metadata |> filter(!Subject_ID %in% setdiff(metadata$Subject_ID, colnames(abundance)))

```

### Creating a taxonomic annotation table

We create the taxonomy data in the next few steps.

::: {.callout-note}
We see here that we do not have the complete taxonomy lineage for the features in the abundance data, but only the last available taxonomic level (e.g., species). This will make it difficult to filter the dataset to keep only bacterial species, since we do not have the taxonomic annotation for the kingdom level. *Can we assume that all species are bacteria and go ahead with the analyses?*
:::

We first create the taxonomy table by extracting the taxonomic annotation from the feature names in the abundance data, and by keeping only the last available level for each feature, saved as `taxon_names`.

```{r}

taxonomy <- 
  tibble(full_name =  data[1] |> pull()) |>
  filter(!str_detect(full_name, "-1")) |>
  mutate(
    taxon_name = full_name,
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = str_c("g_", word(taxon_name, 1)),
    species = case_when(
      word(taxon_name, 2) == "sp." ~ NA,
      is.na(word(taxon_name, 2)) ~ NA,
      TRUE ~ str_c("s_", word(taxon_name, 2)),
    ),
    taxon = taxon_name
  ) |>
  select(kingdom, phylum, class, order, family, genus, species, taxon_name, taxon)

```

We check if there are duplicates in taxa species, and document them below.

```{r}
  
cat("There are", nrow(taxonomy |> group_by(taxon_name) |> summarise(n = n()) |> filter(n > 1)), 
    "duplicated taxonomic annotations.\n")

```

No duplicates are found in the taxonomic annotation, so we can keep all features in the abundance data, and we can simply set the taxonomic annotation as row names for the taxonomy data.

```{r}

taxonomy <- taxonomy |> column_to_rownames("taxon")

taxonomy |> glimpse()

```

For abundance data, we only need to set the taxonomic annotation as row names and filter the dataset to keep only taxa with taxonomic annotation in the taxonomy data.

```{r}

abundance <- data |> column_to_rownames(var = names(data)[1])

abundance <- abundance |> filter(rownames(abundance) %in% rownames(taxonomy))

```


### Documenting available metadata

For sample (metadata) data, we only keep relevant metadata variables for our analyses, and we set the sample IDs as row names. 

We first document the original metadata variables, and then we select the relevant ones for our analyses.

```{r}

samples <- metadata |> clean_names()

```

::: panel-tabset
#### Age
```{r}
#| results: asis
samples |>
  mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ "H"
  )) |>
  ggplot() +
  aes(x = cohort, y = age, color = cohort, fill = cohort) +
  geom_violin(alpha = 0.6, color = NA) +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.8) +
  labs(x = "", y = "Age", color = "Group", fill = "Group",
       subtitle = "Distribution of age across groups") +
  theme(
    axis.text.x = element_blank()
  )
```

#### BMI
```{r}
#| results: asis
samples |>
  mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ "H"
  )) |>
  mutate(
    bmi = case_when(
      bmi == "NA" ~ NA_real_,
      TRUE ~ as.numeric(bmi)
    )
  ) |>
  ggplot() +
  aes(x = cohort, y = bmi, color = cohort, fill = cohort) +
  geom_violin(alpha = 0.6, color = NA) +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.8) +
  labs(x = "", y = "BMI", color = "Group", fill = "Group",
       subtitle = "Distribution of BMI across groups") +
  theme(
    axis.text.x = element_blank()
  )
```

#### Gender
```{r}
#| results: asis
samples |>
   mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ "H"
  )) |>
  ggplot() +
  aes(x = gender, fill = cohort) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Count",fill = "Group",
       subtitle = "Distribution of gender across groups")
```

#### Group
```{r}
#| results: asis
  samples |>
   mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ "H"
  )) |>
  ggplot() +
  aes(x = cohort, fill = cohort) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Count", fill = "Group",
       subtitle = "Distribution of samples across groups") +
  theme(
    axis.text.x = element_blank()
  )
```
:::

We document the different levels of the variable `Group` below.

```{r}

metadata |> 
  clean_names() |> 
  select(subject_id, group) |>
  group_by(group) |> 
  summarise(n = n()) |>
  gt(caption = "Health status of samples")

```


The selected variables are:

- `Subject_ID`: the sample ID, which will be set as row names.
- `Group`: the health status of the samples.

We filter the metadata to keep only the selected variables, set the sample IDs as row names, and document all the variables in a dictionary, documenting the original variable name, the new variable name, the description of the variable, the values it can take, and if it is kept for further analyses or not.

```{r}

samples <- 
  metadata |> 
  clean_names() |> 
  select(subject_id, group) |>
  column_to_rownames("subject_id") |>
  mutate(group = group |> as.factor()) |>
  mutate(cohort = case_when(
    group == "Healthy" ~ "H",
    group %in% c("MP", "Stage_0", "Stage_I_II", "Stage_III_IV") ~ "CRC",
    group == "HS" ~ NA_character_
  ))

```

We filter samples for which we do not have a clear health status (i.e., samples with "HS" group).

```{r}

samples <- samples |> filter(!is.na(cohort)) 

samples |> glimpse()

```

We also filter abundance data to keep only samples for which we have a clear health status in the metadata.

```{r}

abundance <- abundance |> select(any_of(rownames(samples)))

abundance |> head() |> select(1:5)

```

#### Creating a dictionary for metadata variables

> TODO: add dictionnary

### Documenting abundance distribution

We check if there are any singletons (i.e., taxa that are present in only one sample) in the dataset, and if so, we document them by showing the taxonomic annotation and the number of samples with this annotation.

```{r}

abundance |> 
  rownames_to_column("taxon") |>
  pivot_longer(-taxon, names_to = "sample_id", values_to = "rel_ab") |>
  filter(rel_ab > 0.05) |>
  group_by(taxon) |>
  mutate(singleton = ifelse(n() == 1, "Singleton", "Non-singleton")) |>
  ungroup() |>
  filter(singleton == "Singleton") |>
  ggplot() +
  aes(x = taxon, y = sample_id, fill = rel_ab) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue", labels = scales::label_percent(accuracy = 1)) +
  labs(x = "Taxon", y = "Sample ID", fill = "Relative abundance",
       subtitle = "Singleton taxa where rel_ab > 5% (present in only one sample)") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  mutate(rel_ab = count / sum(count)) |>
  ungroup() |>
  filter(rel_ab > 0) |>
  group_by(sample_id) |>
  summarise(n_taxa = n()) |>
  ggplot() +
  aes(x = n_taxa) +
  geom_histogram(bins = 30) +
  labs(x = "Number of taxa", y = "Number of samples",
       subtitle = "Distribution of the number of taxa per sample") +

abundance |> 
  rownames_to_column("sample_id") |>
  pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
  group_by(sample_id) |>
  summarise(mean_count = mean(count)) |>
  ggplot() +
  aes(x = mean_count |> log10()) +
  geom_histogram(bins = 30) +
  labs(x = "Mean counts", y = "Number of samples",
       subtitle = "Distribution of the mean counts across samples") +

  plot_layout(nrow = 2)

```

### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_crc <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_crc

```


### CDI dataset (Ferretti et al., 2023)

The *Clostridium difficile* infection (CDI) metagenomics dataset was obtained from the [Github repository](https://github.com/grp-bork/cdifficile_Ferretti_2022/tree/main/data/CDI) associated with the original article from Ferretti et al. (2023). [indicate the date of the latest version for these datasets]. 


For the analysis, we have the following datasets at our disposal: 

- `CDI_mergedControl_samples.csv`: contains metadata for both control and CDI samples, in CSV format.
- `mOTU20_profiles_filteredCDI.csv`: contains abundance count data for stool samples, in CSV format.
- `motus_taxonomy.csv`: contains the taxonomic annotation for the features in the abundance data, in CSV format.

### Importing raw data

In the chunks below, we will import the metadata, abundance data, and taxonomic annotation for the CDI dataset.

```{r}

metadata <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "CDI_mergedControl_samples.csv"))
abundance <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "mOTU20_profiles_filteredCDI.csv"))
taxonomy <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "motus_taxonomy.csv"))

```

::: {.callout-note}
Interestingly, this dataset contains samples from different study (and thus different sequencing protocols).
:::

#### Documenting the substudies protocols from which the samples were obtained

We document below the number of samples in each of the different studies included in the dataset.

```{r}

metadata |>
  janitor::clean_names() |>
  group_by(study) |>
  summarise(n = n()) |>
  gt(caption = "Number of samples per study")

```


Below, we document the study protocol for each of the studies included in the dataset, by looking at the original articles associated with each of the studies.

* [Kim et al., 2020 (Nature scientific reports)](https://www.nature.com/articles/s41598-020-74090-0): through July 2016 to June 2018, individuals diagnosed with CDI at Hanyang University Seoul Hospital were screened, and CDI patients who could provide over 50 g of faeces sample from CDI diagnosed stool sample were enrolled in the study.

* [Kumar et al., 2017 (Nature njp biofilms and microbiome)](https://www.nature.com/articles/s41522-017-0020-7): to complete

> TODO: complete this section

::: {.callout-note}
We see here that we do not have the complete taxonomy lineage for the features in the abundance data, but only the last available taxonomic level (e.g., species). This will make it difficult to filter the dataset to keep only bacterial species, since we do not have the taxonomic annotation for the kingdom level. *Can we assume that all species are bacteria and go ahead with the analyses?*
:::

### Creating a taxonomic annotation

First, we need to filter taxonomy data to merge taxon with the same taxonomic annotation, and to keep only one row for each taxon, with the taxonomic annotation as row names. Additionally, we fill the taxonomic levels based on the taxonomic annotation.

```{r}

taxonomy <- 
  taxonomy |>
  mutate(
    taxon_name = coalesce(
      str_remove(mOTU_name, "\\[.*\\]")),
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = str_c("g_", word(taxon_name, 1)),
    species = case_when(
      word(taxon_name, 2) == "sp." ~ NA,
      is.na(word(taxon_name, 2)) ~ NA,
      TRUE ~ str_c("s_", word(taxon_name, 2))
    ),
    taxon = str_c(taxon_name)
  ) |>
  select(kingdom, phylum, class, order, family, genus, species, taxon, taxon_name)

```

We now filter the abundance data by first replacing the feature names in the abundance data with the taxonomic annotation from the taxonomy data, and then summing the counts of features with the same taxonomic annotation to get the count of the taxon.

```{r}

abundance <-
  abundance |>
  filter(X1 != "-1") |>
  column_to_rownames(var = names(abundance)[1]) |>
  as.data.frame() |>
  bind_cols(taxonomy |> select(taxon_name)) |>
  group_by(taxon_name) |>
  summarise(across(everything(), sum)) |>
  column_to_rownames("taxon_name") |>
  as.data.frame()

```

We filter the taxonomy data to keep only one row for each taxon, and to set the taxon names as row names.

```{r}

taxonomy <-
  taxonomy |>
  distinct(taxon_name, .keep_all = TRUE) |>
  column_to_rownames("taxon")

taxonomy |> glimpse()

```

### Documenting available metadata

We document the metadata variables to select the relevant ones for our analyses.

::: panel-tabset

#### Age group
```{r}
#| results: asis
metadata |>
  clean_names() |>
  ggplot() +
  aes(x = age_group) +
  geom_bar() +
  labs(x = "Age group", y = "Count", fill = "Age group",
       subtitle = "Distribution of samples across age groups",
       caption = "Only several studies have age group information, thus is not representative of the whole dataset.")
```

#### Geography
```{r}
#| results: asis
metadata |>
  clean_names() |>
  ggplot() +
  aes(x = geography) +
  geom_bar() +
  labs(x = "Geography", y = "Count", fill = "Geography", 
       subtitle = "Distribution of samples across geographies")
```

#### Antibiotics
```{r}
#| results: asis
metadata |>
  clean_names() |>
  ggplot() +
  aes(x = abx) +
  geom_bar() +
  labs(x = "Antibiotics", y = "Count", fill = "Antibiotics",
       subtitle = "Distribution of samples across antibiotics usage")
```

#### CDI diagnosis
```{r}
#| results: asis
metadata |>
  clean_names() |>
  ggplot() +
  aes(x = cdi) +
  geom_bar() +
  labs(x = "CDI diagnosis", y = "Count", fill = "CDI diagnosis",
       subtitle = "Distribution of samples across CDI diagnosis")

```

#### C.difficile presence
```{r}
#| results: asis
metadata |>
  clean_names() |>
  ggplot() +
  aes(x = cdiff, fill = cdi) +
  geom_bar() +
  labs(x = "C.Difficile presence", y = "Count", fill = "CDI diagnosed",
       subtitle = "Distribution of samples across C. difficile presence")
```
:::


#### Documenting raw metadata variables in a dictionary

> TO DO: Create a dictionary to document original variable names, new variable names, description of the variables, values they can take, and if they are kept for further analyses or not.

We will now filter the metadata to keep only relevant variables for our analyses, and to set the sample IDs as row names. The selected variables are: 

- `Samples`: the sample ID, which will be set as row names.
- `Gut_sample_type`: the type of the sample, here only "stool" samples are available.
- `Status`: the health status of the sample, with two levels: "Healthy" and "Diseased".
- `subject_disease_status`: gives further information about the health status of the samples.

```{r}

samples <-
  metadata |>
  clean_names() |>
  select(samples, gut_sample_type, status, subject_disease_status) |>
  column_to_rownames("samples")

```

#### Filtering disease status to keep only CDI samples

We document the variable `subject_disease_status` to see how many samples have been diagnosed with other gut-related diseases.

```{r}

samples |>
  group_by(status) |>
  summarise(n = n()) |>
  gt(caption = "Number of healthy and diseased samples")

```


```{r}

samples |>
  group_by(status, subject_disease_status) |>
  summarise(n = n()) |>
  gt(caption = "Health status of healthy and diseased samples")

```

::: {.callout-warning}
Based on the documentation of the variable `subject_disease_status`, we see that all of the "Healthy" samples are control and none of them have a gut-related disease. However, among "Diseased" samples, only 202 among the 348 diseased samples have been diagnosed with CDI. *Do we keep all the "Diseased" samples for our analyses, or do we only keep the 202 samples diagnosed with CDI?*.
:::

I filter the dataset to keep only the 202 samples diagnosed with CDI, as well as the healthy samples (CTR).

::: {.callout-note}
I could have also kept some other diseased samples, such as those "Asymptotically colonized", but there are only four of them.
:::

```{r}

# samples <- samples |> filter(subject_disease_status == "CDI" | subject_disease_status == "CTR")

```

We filter the abundance data to keep only the samples for which we have the corresponding filtered metadata.

```{r}

# abundance <- abundance |>
#   select(any_of(rownames(samples)))

```


### Rarefaction and filtering of low-abundance taxa

We will then rarefy the CDI dataset, so that all samples have the same sequencing depth, and filter out low-abundance taxa. We will use the `rrarefy` function from the `vegan` package to perform rarefaction.

First, we separate the dataset into two groups: "healthy" and "diseased" so that rarefaction (and prevalence filtering) is performed separately for the two groups, to avoid biasing the results by removing more samples from one group than the other.

```{r}

tmp <-
  abundance |>
  rownames_to_column("taxon") |>
  pivot_longer(-taxon, names_to = "sample_id", values_to = "count") |>
  left_join(
    samples |> rownames_to_column("sample_id") |> select(sample_id, status),
    by = join_by(sample_id)
  )

```


```{r}

tmp_healthy <- tmp |> filter(status == "Healthy")
tmp_diseased <- tmp |> filter(status == "Diseased")

```

We show the distribution of sequencing depth across samples, to choose a rarefaction depth.

::: {.callout-note}
Note that a classical rarefaction threshold would be to simply take the median depth, but in this case, we choose to take the median depth divided by 2, to keep more samples in the dataset.
:::

```{r}

tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across healthy samples")

tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count)) |>
  ggplot(aes(x = depth)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(depth)/2), color = "red", linetype = "dashed") +
  labs(x = "Sequencing depth", y = "Number of samples", caption = "Red dashed line represent median depth divided by 2",
       subtitle = "Distribution of sequencing depth across diseased samples")

```

We will rarefy at a depth of median(depth)/2; this is done in the chucks below, where we also filters out samples with depth < median(depth)/2.

```{r}

# for healthy samples

tmp_depth <- tmp_healthy |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2 |> round()

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_healthy <- tmp_healthy |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```


```{r}

# for diseased samples

tmp_depth <- tmp_diseased |>
  group_by(sample_id) |>
  summarise(depth = sum(count), .groups = "drop")

min_depth <- median(tmp_depth$depth) / 2 |> round()

samples_to_keep <- tmp_depth |>
  filter(depth >= min_depth) |>
  pull(sample_id)

cat("Removing", sum(tmp_depth$depth < min_depth), "samples below depth", round(min_depth))

tmp_rarefied_diseased <- tmp_diseased |>
    filter(sample_id %in% samples_to_keep) |>
    select(sample_id, taxon, count) |>
    pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
    column_to_rownames("sample_id") |>
    rrarefy(sample = min_depth) |>
    as.data.frame() |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "taxon", values_to = "count") |>
    group_by(sample_id) |>
    mutate(rel_ab = count / sum(count)) |>
    ungroup()

```

We convert the rarefied data back to wide format, and we merge the healthy and diseased datasets back together. Additionally, we filter the abundance data to only keep taxa for which the total count across samples is greater than 0, to remove taxa that are not present in any sample after rarefaction.

```{r}

abundance <- bind_rows(tmp_rarefied_healthy, tmp_rarefied_diseased) |>
  select(sample_id, taxon, count) |>
  pivot_wider(names_from = taxon, values_from = count, values_fill = 0) |>
  column_to_rownames("sample_id") |>
  as.data.frame() |>
  select(where(~ sum(.) > 0))

abundance |> head() |> select(1:5)

```

Since several samples have been removed after rarefaction, we also need to filter the metadata to keep only the samples for which we have abundance data after rarefaction.

> TODO: Document the taxa for which there are no more counts after rarefaction, and check for other datasets as well.

```{r}

samples <- samples |> filter(rownames(samples) %in% rownames(abundance))

```

We also filter the taxonomy table to keep only taxa for which we have abundance data after rarefaction.

```{r}

taxonomy <- taxonomy |> filter(rownames(taxonomy) %in% colnames(abundance))

```

### Saving cleaned data into a `phyloseq` object

We finally create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_cdi <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_cdi

```


```{r}
#| include: false

rm(data, data_biopsy, metadata, df, samples, taxonomy, abundance)

```

