---
title: "Data preprocessing for dysbiotic cohorts"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: true
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

In this section, we will first import the relevant datasets from the dysbiotic cohorts (that is, those pertaining to IBD, IBS, CRC, and CDI), and then perform several descriptive analyses for each of those datasets. The analyses will include: data cleaning, taxonomic aggregation (if necessary) and rarefaction. 

We will transform each dataset into a `phyloseq` object, before merging them into a `MultiAssayExperiment` object in the last section. For this we need:

- abundance data, a matrix of relative abundance with samples as rows and features (taxa) as columns.
- taxonomy data, a data frame with features (taxa) as rows and taxonomic levels as columns.
- samples data, a data frame with samples as rows and metadata variables as columns.

> TODO: Create dictionaries for metadata variables for each dataset, to keep track of the meaning of each variable and the values they can take, previous and new name, and if it is kept for further analysis or not.

## IBD dataset (Lloyd-Price et al. 2019)

The inflammatory bowel disease (IBD) dataset was obtained from the [IBDMDB](https://ibdmdb.org/), with the results published in Lloyd-Price et al. (2019). The authors of the Science paper did not provide information to which study name the IBD dataset comes from, so the HMP2 study was chosen, which seems like a reasonable choice since it is the most recent study on this page. 

Stool samples were initially collected from 132 participants at multiple time points. Specifically, samples were collected at baseline (that is after the initial colonoscopy, which was used to stratify participants into healthy controls, Crohn's disease (CD) patients, and ulcerative colitis (UC) patients), and then every 2 weeks for 1 years.

Metagenomics (MGX) were performed on 1638 stool samples; DNA was quantified, normalized, and used to prepare sequencing libraries with the Nextera XT kit. Libraries were pooled, quality-checked, and sequenced on an Illumina HiSeq platform to generate ~10 million paired-end reads per sample. Raw sequencing data was then demultiplexed and converted to standard file formats (BAM and FASTQ files) using the Picard suite.

### Importing raw data

In the chunks below, we import the metadata and species count data for the IBD dataset.

> TODO: Document the names of the files they imported in their analysis

```{r}

metadata <- read_tsv(file.path(data_dir(), "01-raw-data/01bis-IBD", "metadata.tsv"))
abundance <- read_tsv(file.path(data_dir(), "01-raw-data/01bis-IBD", "species.counts.tsv"))

```

There are `r nrow(metadata)` samples and `r ncol(abundance)` features in the original dataset.

#### Reorganizing the taxonomy in the abundance data and creating a taxonomy table

```{r}

abundance <- abundance |> column_to_rownames("Sample")

```


```{r}

taxonomy <- 
  tibble(full_name = names(abundance)) |>
  mutate(
    domain = str_extract(full_name, "d__[^;]+"),
    phylum = str_extract(full_name, "p__[^;]+"),
    class = str_extract(full_name, "c__[^;]+"),
    order = str_extract(full_name, "o__[^;]+"),
    family = str_extract(full_name, "f__[^;]+"),
    genus = str_extract(full_name, "g__[^;]+"),
    species = str_extract(full_name, "s__[^;]+")
  ) |>
  mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, domain
    )
  ) |>
  mutate(
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(domain, "d__"), " (d)")
    )
  ) |>
  filter(domain == "d__Bacteria") |>
  select(-full_name)

```


```{r}
# Checking "noname" and "unclassified" annotations in the taxonomy data (table)

taxonomy |>
  summarise(across(c(domain, phylum, class, order, family, genus, species), list(
    unclassified = ~ sum(str_detect(., "_unclassified"), na.rm = TRUE),
    noname = ~ sum(str_detect(., "_noname"), na.rm = TRUE)
  ))) |>
  pivot_longer(everything(), names_to = c("taxonomic_level", "annotation"), names_sep = "_") |>
  pivot_wider(names_from = annotation, values_from = value) |>
  gt(caption = "Number of unclassified and noname annotations across taxonomic levels in the taxonomy data")

```

::: {.callout-note}
Contrary to the IBD dataset from the IBDMBD website, this dataset does not contain any "unclassified" or "noname" annotations in the taxonomy data.
:::

```{r}
# replacing cleaned names in abundance table

abundance <- abundance |> set_names(taxonomy$taxon_name)

# set taxon names as row names in taxonomy table

taxonomy <- taxonomy |> column_to_rownames("taxon_name")

```

#### Documenting the metadata variables

```{r}
samples <- metadata |> clean_names()
samples |> glimpse()
```
```{r}

# set sample names as row names in metadata table

samples <- samples |> column_to_rownames("sample")

```


### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibd_alt <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibd_alt

```

## IBS dataset (Mars et al., 2020)

The irritable bowel syndrome (IBS) dataset in Mars et al. (2020) was obtained from [Mendeley data](https://data.mendeley.com/datasets/29n2z5r5ph/3) plateform. 

Metagenomics were performed for 474 stool samples recolted from 77 participants, separated into healthy controls (H),  constipation predominant (IBS-C), diarrhea predominant (IBS-D), or mixed (IBS-M). DNA was extracted from stool samples using the QIAGEN PowerSoil kit and quantified by NanoDrop and PicoGreen assays. Shotgun metagenomic libraries were prepared using a modified Nextera XT protocol and sequenced on either a HiSeq 2500 or NextSeq platform with single-end reads. Samples from three cohorts were randomly distributed across two sequencing runs to reduce batch effects. Reads were trimmed to 100 bp and aligned to a prokaryotic reference genome collection at 97% identity using BURST, with low-depth samples (< 10,000 reads) removed. Functional profiles were generated using KEGG Orthology annotations, and low-abundance gene estimates were further improved using SHOGUN.

### Importing raw data

```{r}

metadata <- read_tsv(file.path(data_dir(), "01-raw-data/02bis-IBS", "metadata.tsv"))
abundance <- read_tsv(file.path(data_dir(), "01-raw-data/02bis-IBS", "species.counts.tsv"))

```

#### Reorganizing the taxonomy in the abundance data and creating a taxonomy table

```{r}

abundance <- abundance |> column_to_rownames("Sample")

```

```{r}

taxonomy <- 
  tibble(full_name = names(abundance)) |>
  mutate(
    domain = str_extract(full_name, "d__[^;]+"),
    phylum = str_extract(full_name, "p__[^;]+"),
    class = str_extract(full_name, "c__[^;]+"),
    order = str_extract(full_name, "o__[^;]+"),
    family = str_extract(full_name, "f__[^;]+"),
    genus = str_extract(full_name, "g__[^;]+"),
    species = str_extract(full_name, "s__[^;]+")
  ) |>
  mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, domain
    )
  ) |>
  mutate(
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(domain, "d__"), " (d)")
    )
  ) |>
  filter(domain == "d__Bacteria") |>
  select(-full_name)

```

Same as the previous dataset, there are no "unclassified" or "noname" annotations in the taxonomy data.

```{r}
# replacing cleaned names in abundance table

abundance <- abundance |> set_names(taxonomy$taxon_name)

# set taxon names as row names in taxonomy table

taxonomy <- taxonomy |> column_to_rownames("taxon_name")

```

#### Documenting the metadata variables

```{r}
samples <- metadata |> clean_names()
samples |> glimpse()
```

```{r}
# set sample names as row names in metadata table

samples <- samples |> column_to_rownames("unique_id")

# modify abundance table to have the same sample names as in the metadata table

abundance <- abundance |> magrittr::set_rownames(rownames(samples))

```


### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_ibs_alt <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_ibs_alt

```

## CRC dataset (Yachida et al., 2019)

The colorectal cancer (CRC) dataset in Yashida et al. (2019) was obtained from the DDBJ Sequence Read Archive (DRA) under the names [DRA006684](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA006684) and [DRA008156](https://ddbj.nig.ac.jp/search/entry/sra-submission/DRA008156); however, those are raw sequencing data, and the authors mentioned having used only processed datasets.

Processed datasets seems to be available as [Supplementary Tables](https://www.nature.com/articles/s41591-019-0458-7#Sec23), therefore these will be used for the subsequent analyses.

Stool samples were collected after bowel prep and stored on dry ice. DNA was extracted using bead-beating with a GNOME kit, quality-checked by TapeStation, and stored at −80°C. Sequencing libraries were prepared with the Nextera XT kit and verified for quality before paired-end whole-genome shotgun sequencing on an Illumina HiSeq 2500, generating 150 bp reads with a target output of 5 Gb per sample.

The samples were classified into nine groups according to colonoscopic and histological findings: 

- normal (no remarkable colonoscopic findings) and a few polyps (up to two small (<5 mm) polyps) were classified as "Healthy";
- multiple polypoid adenomas with low-grade dysplasia and polypoid adenoma(s) with high-grade dysplasia were classified as "MP";
- stage 0 CRC was classified as "Stage_0"; 
- stage I CRC and stage II CRC were classified as "Stage_I_II"; 
- stage III CRC and stage IV CRC were classified as "Stage_III_IV". 
- the remaining group was normal with a history of colorectal surgery, and was classified as "HS".


### Importing raw data

```{r}

metadata <- read_tsv(file.path(data_dir(), "01-raw-data/03bis-CRC", "metadata.tsv"))
abundance <- read_tsv(file.path(data_dir(), "01-raw-data/03bis-CRC", "species.counts.tsv"))

```

#### Reorganizing the taxonomy in the abundance data and creating a taxonomy table

```{r}

taxonomy <-
  tibble(full_name = names(abundance)[-1]) |>
  mutate(
    domain = str_extract(full_name, "d__[^;]+"),
    phylum = str_extract(full_name, "p__[^;]+"),
    class = str_extract(full_name, "c__[^;]+"),
    order = str_extract(full_name, "o__[^;]+"),
    family = str_extract(full_name, "f__[^;]+"),
    genus = str_extract(full_name, "g__[^;]+"),
    species = str_extract(full_name, "s__[^;]+")
  ) |>
  mutate(
    taxon = coalesce(
      species, genus, family, order, class, phylum, domain
    )
  ) |>
  mutate(
    taxon_name = coalesce(
      str_remove(species, "s__"),
      str_c(str_remove(genus, "g__"), " (g)"),
      str_c(str_remove(family, "f__"), " (f)"),
      str_c(str_remove(order, "o__"), " (o)"),
      str_c(str_remove(class, "c__"), " (c)"),
      str_c(str_remove(phylum, "p__"), " (p)"),
      str_c(str_remove(domain, "d__"), " (d)")
    )
  ) |>
  filter(domain == "d__Bacteria") |>
  select(-full_name)

```

```{r}
# replacing cleaned names in abundance table

abundance <- abundance |> set_names(c("sample_id", taxonomy$taxon_name)) |> column_to_rownames("sample_id")

# set taxon names as row names in taxonomy table

taxonomy <- taxonomy |> column_to_rownames("taxon_name")

```

#### Documenting the metadata variables

```{r}

samples <- metadata |> clean_names()
samples |> glimpse()

```

```{r}
# set sample names as row names in metadata table

samples <- samples |> column_to_rownames("sample")

```


### Saving cleaned data into a `phyloseq` object

Finally, we can create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_crc_alt <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_crc_alt

```

## CDI dataset (Ferretti et al., 2023)

The *Clostridium difficile* infection (CDI) metagenomics dataset was obtained from the [Github repository](https://github.com/grp-bork/cdifficile_Ferretti_2022/tree/main/data/CDI) associated with the original article from Ferretti et al. (2023). [indicate the date of the latest version for these datasets]. 

For the analysis, we have the following datasets at our disposal: 

- `CDI_mergedControl_samples.csv`: contains metadata for both control and CDI samples, in CSV format.
- `mOTU20_profiles_filteredCDI.csv`: contains abundance count data for stool samples, in CSV format.
- `motus_taxonomy.csv`: contains the taxonomic annotation for the features in the abundance data, in CSV format.

::: {.callout-note}
For the original analysis, the authors filtered out -1 column, column containing "unknown" annotation, and columns with meta-mOTU, which represent metagenomics mOTU.
:::

### Importing raw data

```{r}

metadata <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "CDI_mergedControl_samples.csv"))
abundance <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "mOTU20_profiles_filteredCDI.csv"))
taxonomy <- read_csv(file.path(data_dir(), "01-raw-data/04-CDI", "motus_taxonomy.csv"))

```

#### Documenting the substudies protocols from which the samples were obtained

```{r}

metadata |>
  janitor::clean_names() |>
  group_by(study) |>
  summarise(n = n()) |>
  gt(caption = "Number of samples per study")

```

#### Filtering `unknown` and `meta-mOTU` annotations in the abundance data

```{r}

abundance <- abundance |> column_to_rownames("X1")

```

```{r}

abundance <- abundance |> dplyr::filter_out(str_detect(rownames(abundance), "meta_mOTU|unknown|-1"))
  
```

#### Creating a taxonomic annotation table from taxonomy data

```{r}
# filter taxonomy table to only keep selected OTU in abundance

taxonomy <- taxonomy |> filter(mOTU_name %in% rownames(abundance))

```

```{r}

setdiff(rownames(abundance), taxonomy$mOTU_name)

```

We remove this feature from the abundance table, since it does not have a taxonomic annotation in the taxonomy table.

```{r}

abundance <- abundance |> dplyr::filter_out(str_detect(rownames(abundance),"ref_mOTU_v2_2364"))

```


```{r}

taxonomy <- 
  taxonomy |>
  mutate(
    taxon_name = coalesce(
      str_remove(mOTU_name, "\\[.*\\]")),
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = str_c("g_", word(taxon_name, 1)),
    species = case_when(
      word(taxon_name, 2) == "sp." ~ NA,
      is.na(word(taxon_name, 2)) ~ NA,
      TRUE ~ str_c("s_", word(taxon_name, 2))
    ),
    strain = case_when(
      word(taxon_name, 2) == "sp." & !is.na(word(taxon_name, 3)) & word(taxon_name, 3) != "" ~ str_c("st_", word(taxon_name, 3)),
      TRUE ~ NA_character_
    ),
    taxon = str_c(taxon_name)
  ) |>
  select(kingdom, phylum, class, order, family, genus, species, strain, taxon, taxon_name, mOTU_name)


taxonomy <- 
  taxonomy |>
  mutate(
    taxon_name = case_when(
      !is.na(strain) ~ str_c(taxon, "(strain)"),
      !is.na(species) ~ str_c(taxon),
      !is.na(genus) ~ str_c(taxon, " (genus)"),
      !is.na(family) ~ str_c(taxon, " (family)"),
      !is.na(order) ~ str_c(taxon, " (order)"),
      !is.na(class) ~ str_c(taxon, " (class)"),
      !is.na(phylum) ~ str_c(taxon, " (phylum)"),
      TRUE ~ str_c(taxon, " (kingdom)")
    )
  )
  
```


```{r}
# filter abundance table and sum counts for features with the same taxonomic annotation

abundance <- abundance |> 
  rownames_to_column("mOTU_name") |>
  left_join(taxonomy |> select(mOTU_name, taxon_name), by = "mOTU_name") |>
  group_by(taxon_name) |>
  summarise(across(-mOTU_name, sum)) |>
  column_to_rownames("taxon_name")

```

```{r}

# we filter taxonomy table to only keep one row per taxonomic annotation

taxonomy <- taxonomy |> group_by(taxon_name) |> dplyr::slice(1) |> ungroup() |> column_to_rownames("taxon_name")

```

#### Documenting the metadata variables

```{r}

samples <- metadata |> clean_names()
samples |> glimpse()

```

```{r}
# set sample names as row names in metadata table

samples <- samples |> column_to_rownames("samples")

```

### Saving cleaned data into a `phyloseq` object

We finally create the `phyloseq` object using the `otu_table`, `tax_table`, and `sample_data` functions from the `phyloseq` package.

```{r}

ps_cdi_alt <- phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxonomy)),
  sample_data(samples)
)

ps_cdi_alt

```

```{r}

rm(abundance, metadata, taxonomy, samples)

```


