---
title: "Descriptive analysis for supp. cohorts"
author:
  - Mariam Rahali Masbouri
date: today 
format:
  html:
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    number-sections: true
    embed-resources: true
execute:
  cache: false 
  warning: false
project:
  execute-dir: project
editor: source
editor_options: 
  chunk_output_type: inline
---

## IBD dataset from Fransoza et al. (2019)

### Importing raw data

```{r}

data <- 
  read_xlsx(file.path(data_dir(), "01-raw-data/07-IBD-Franzosa", "41564_2018_306_MOESM6_ESM.xlsx"), skip = 1) |> 
  clean_names() |> 
  column_to_rownames("number_feature_sample")

```

### Constructing metadata table

```{r}

metadata <- 
  data |> 
  t() |> 
  as_tibble(rownames = "uid") |> 
  select(1:9) |> 
  column_to_rownames("uid") |> 
  clean_names()

metadata |> glimpse()

```

We document below the number of samples per diagnosis group in the IBD dataset from Fransoza et al. (2019).

```{r}

metadata |>
  group_by(diagnosis) |>
  summarise(n = n()) |>
  ungroup() |>
  gt(caption = "Number of samples per diagnosis group in the IBD dataset from Fransoza et al. (2019)")

```


### Constructing abundance table

```{r}

abundance <- 
  data |> 
  as_tibble(rownames = "uid") |> 
  slice_tail(n = 201) |>
  column_to_rownames("uid") |>
  mutate(across(everything(), as.numeric)) |>
  t()

```

### Constructing taxonomy table based on abudance data

Since no taxonomy table is provided in the original dataset, we will construct a taxonomy table based on the abundance data.

```{r}

taxonomy <-
  tibble(
    taxon = colnames(abundance),
    kingdom = "k_Bacteria",
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = word(taxon, 1, sep = "_"),
    species = word(taxon, 2, sep = "_")
  ) |>
  mutate(
    species = case_when(
      species == "sp" ~ NA_character_,
      TRUE ~ species
    )
  ) |>
  column_to_rownames("taxon")

```

### Creating a `phyloseq` object and saving it for later use

```{r}

ps_ibd_franzosa <- 
  phyloseq(
  otu_table(as.matrix(abundance), taxa_are_rows = FALSE),
  sample_data(metadata),
  tax_table(as.matrix(taxonomy))
)

ps_ibd_franzosa

```

## CRC dataset from Zeller et al. (2014)

### Importing raw data

```{r}

data <- curatedMetagenomicData("ZellerG_2014.relative_abundance", dryrun = FALSE, counts = TRUE)[[1]]

```

### Extracting abundance data from the MAE object

The abundance matrix is a count matrix because we set `counts = TRUE` in the `curatedMetagenomicData` function.

```{r}

abundance <- assay(data, "relative_abundance")

```

### Extracting taxonomy data from the MAE object


```{r}

taxonomy <- 
  data |> 
  rowLinks() |>
  as_tibble() |>
  clean_names() |>
  select(taxon_name = node_lab) |>
  separate(taxon_name, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "\\|") |>
  mutate(taxon = str_remove(species, "s__")) |>
  column_to_rownames("taxon")

```

We use taxonomy rownames and replace rownames in the abundance matrix.

```{r}

abundance <- abundance |> magrittr::set_rownames(rownames(taxonomy))

```

### Extracting metadata from the MAE object

```{r}

metadata <- 
  data |> 
  colData() |>
  as_tibble(rownames = "uid") |>
  clean_names() |>
  column_to_rownames("uid")

metadata |> glimpse()

```

We document below the number of samples per diagnosis group in the CRC dataset from Zeller et al. (2014).

```{r}

metadata |>
  group_by(disease, study_condition) |>
  summarise(n = n()) |>
  ungroup() |>
  gt(caption = "Number of samples per diagnosis group in the CRC dataset from Zeller et al. (2014)")

```


### Creating a `phyloseq` object and saving it for later use

```{r}

ps_crc_zeller <- 
  phyloseq(
    otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
    sample_data(metadata),
    tax_table(as.matrix(taxonomy))
  )

ps_crc_zeller

```


## CDI dataset from Vincent et al. (2016)

### Importing raw dataset

```{r}

data <- curatedMetagenomicData("VincentC_2016.relative_abundance", dryrun = FALSE, counts = TRUE)[[1]]

```

### Extracting abundance data from the MAE object

```{r}

abundance <- assay(data, "relative_abundance")

```


### Extracting taxonomy data from the MAE object

```{r}

taxonomy <- 
  data |> 
  rowLinks() |>
  as_tibble() |>
  clean_names() |>
  select(taxon_name = node_lab) |>
  separate(taxon_name, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "\\|") |>
  mutate(taxon = str_remove(species, "s__")) |>
  column_to_rownames("taxon")

```

We use taxonomy rownames and replace rownames in the abundance matrix.

```{r}

abundance <- abundance |> magrittr::set_rownames(rownames(taxonomy))

```


### Extracting metadata from the MAE object

```{r}

metadata <- 
  data |> 
  colData() |>
  as_tibble(rownames = "uid") |>
  clean_names() |>
  column_to_rownames("uid")

metadata |> glimpse()

```

We document below the number of samples per diagnosis group in the CDI dataset from Vincent et al. (2016).

```{r}

metadata |>
  group_by(disease, study_condition) |>
  summarise(n = n()) |>
  ungroup() |>
  gt(caption = "Number of samples per diagnosis group in the CDI dataset from Vincent et al. (2016)")

```


### Creating a `phyloseq` object and saving it for later use

```{r}

ps_cdi_vincent <- 
  phyloseq(
    otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
    sample_data(metadata),
    tax_table(as.matrix(taxonomy))
  )

ps_cdi_vincent

```

## CDI dataset from Ianiro et al. (2022)

### Importing raw dataset

```{r}

data <- curatedMetagenomicData("IaniroG_2022.relative_abundance", dryrun = FALSE, counts = TRUE)[[1]]

```

### Extracting abundance data from the MAE object

```{r}

abundance <- assay(data, "relative_abundance")

```

### Extracting taxonomy data from the MAE object

```{r}

taxonomy <- 
  data |> 
  rowLinks() |>
  as_tibble() |>
  clean_names() |>
  select(taxon_name = node_lab) |>
  separate(taxon_name, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "\\|") |>
  mutate(taxon = str_remove(species, "s__")) |>
  column_to_rownames("taxon")

```

We use taxonomy rownames and replace rownames in the abundance matrix.

```{r}

abundance <- abundance |> magrittr::set_rownames(rownames(taxonomy))

```

### Extracting metadata from the MAE object

```{r}

metadata <- 
  data |> 
  colData() |>
  as_tibble(rownames = "uid") |>
  clean_names() |>
  column_to_rownames("uid")

metadata |> glimpse()

```

```{r}

metadata |>
  group_by(disease, study_condition) |>
  summarise(n = n()) |>
  ungroup() |>
  gt(caption = "Number of samples per diagnosis group in the Ianiro et al. (2022) dataset")

```


### Creating a `phyloseq` object and saving it for later use

```{r}

ps_cdi_ianiro <- 
  phyloseq(
    otu_table(as.matrix(abundance), taxa_are_rows = TRUE),
    sample_data(metadata),
    tax_table(as.matrix(taxonomy))
  )

ps_cdi_ianiro

```

