---
title: "Network inference with FlashWeave"
author:
  - Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

### CRC cohort (alternative analysis)

```{r}

tmp <-
  mae[["IBD"]] |>
  assay() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("uid") |>
  pivot_longer(-uid, names_to = "taxon", values_to = "rel_ab") |>
  left_join(
    mae[["IBD"]] |> colData() |> as_tibble(rownames = "uid") |> select(uid, pid, visit_num, diagnosis),
    by = join_by(uid)
  )

```

```{r}

tmp_healthy <- tmp |> filter(diagnosis == "nonIBD")
tmp_diseased <- tmp |> filter(diagnosis == "UC" | diagnosis == "CD")

# for healthy samples

healthy_df <-
  tmp_healthy |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

# for diseased samples

diseased_df <-
  tmp_diseased |>
  select(uid, taxon, rel_ab) |>
  pivot_wider(names_from = taxon, values_from = rel_ab) |>
  column_to_rownames("uid") |>
  t()

```

#### Running FlashWeave on all taxa

In the chunk code below, we run FlashWeave on all taxa for healthy and diseased samples separately, with the following parameters: `n_bootstrap = 100`, `subsample_pcts = c(0.6, 0.7, 0.8)`, `max_attempts = 120`, and `prevalence_threshold = 0.05`. This means that in each bootstrap iteration, we will subsample the data at different percentages (60%, 70%, and 80%) and run FlashWeave on the subsampled data. The `prevalence_threshold` parameter is set to 0.05, meaning that taxa that are present in less than 5% of the samples will be filtered out before running FlashWeave.

```{r}

# res_healthy <- run_network_bootstrap(healthy_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0.05)
# 
# res_diseased <- run_network_bootstrap(diseased_df, run_fn = run_flashweave, n_bootstrap = 100, subsample_pcts = c(0.6, 0.7, 0.8), max_attempts = 120, prevalence_threshold = 0.05)

```

We save the results in an RDS file for later use.

```{r}

# saveRDS(res_healthy, file = str_c(result_dir(), "/flashweave_res_healthy_IBD_20260328.rds"))
# saveRDS(res_diseased, file = str_c(result_dir(), "/flashweave_res_diseased_IBD_20260328.rds"))

```

The results are imported in the chunk code below.

```{r}

res_healthy <- readRDS(file = str_c(result_dir(), "/flashweave_res_healthy_IBD_20260328.rds"))
res_diseased <- readRDS(file = str_c(result_dir(), "/flashweave_res_diseased_IBD_20260328.rds"))

```

The results are obtained in the chunk code below.

```{r}

df_res_healthy <- get_stats_df(res_healthy, symmetric = TRUE)
df_res_diseased <- get_stats_df(res_diseased, symmetric = TRUE)

df_res <-
  bind_rows(
    df_res_healthy |> mutate(group = "Healthy"),
    df_res_diseased |> mutate(group = "Diseased")
  )

```